\documentclass[10pt]{amsart}

%Packages in use
\usepackage{fullpage, hyperref, vipul, enumerate}

%Title details
\title{Take-home class quiz solutions: Friday January 25: Limits, continuity, differentiation review}
\author{Math 195, Section 59 (Vipul Naik)}
%List of new commands

\begin{document}
\maketitle

{\em Note: Please see somewhat updated/improved language for Q9 -- I
should have stated clearly in the question that you should see options
(D) and (E) before answering. However, doing so would not have
improved your score, as (D) is an incorrect option.}

\section{Performance review}

$24$ people took this $10$ question quiz. The score distribution was
as follows:

\begin{itemize}
\item score of $2$: $2$ people.
\item Score of $3$: $3$ people.
\item Score of $4$: $1$ person.
\item Score of $5$: $2$ people.
\item Score of $6$: $1$ person.
\item Score of $8$: $2$ people.
\item Score of $10$: $13$ people.
\end{itemize}

The question-wise answers and performance summary are below:

\begin{enumerate}
\item Option (B): $17$ people.
\item Option (B): $20$ people.
\item Option (B): $23$ people.
\item Option (D): $14$ people.
\item Option (D): $18$ people.
\item Option (B): $17$ people.
\item Option (B): $21$ people.
\item Option (A): $18$ people.
\item Option (C): $16$ people.
\item Option (A): $15$ people.
\end{enumerate}

\section{Solutions}

\begin{enumerate}
\item Which of the following statements is {\bf always true}? {\em
  Earlier scores: $2/11$, $9/16$, $16/28$}

  \begin{enumerate}[(A)]
  \item The range of a continuous nonconstant function on an open
    bounded interval (i.e., an interval of the form $(a,b)$) is an
    open bounded interval (i.e., an interval of the form $(m,M)$).
  \item The range of a continuous nonconstant function on a closed
    bounded interval (i.e., an interval of the form $[a,b]$) is a
    closed bounded interval (i.e., an interval of the form $[m,M]$).
  \item The range of a continuous nonconstant function on an open
    interval that may be bounded or unbounded (i.e., an interval of
    the form $(a,b)$,$(a,\infty)$, $(-\infty,a)$, or
    $(-\infty,\infty)$), is also an open interval that may be bounded
    or unbounded.
  \item The range of a continuous nonconstant function on a closed
    interval that may be bounded or unbounded (i.e., an interval of
    the form $[a,b]$, $[a,\infty)$, $(-\infty,a]$, or
    $(-\infty,\infty)$) is also a closed interval that may be bounded
    or unbounded.
  \item None of the above.
  \end{enumerate}

  {\em Answer}: Option (B)

  {\em Explanation}: This is a combination of the extreme-value
  theorem and the intermediate-value theorem. By the extreme-value
  theorem, the continuous function attains a minimum value $m$ and a
  maximum value $M$. By the intermediate-value theorem, it attains
  every value between $m$ and $M$. Further, it can attain no other
  values because $m$ is after all the minimum and $M$ the maximum.

  {\em The other choices}:

  Option (A): Think of a function that increases first and then
  decreases. For instance, the function $f(x) := \sqrt{1 - x^2}$ on
  $(-1,1)$ has range $(0,1]$, which is not open. Or, the function
  $\sin x$ on the interval $(0,2\pi)$ has range $[-1,1]$.

  Option (C): The same counterexample as for option (A) works.

  Option (D): We can get counterexamples for unbounded intervals. For
  instance, consider the function $f(x) := 1/x$ on $[1,\infty)$. The
  range of this function is $(0,1]$, which is not closed. The idea is
  that we make the function approach but not reach a finite value as
  $x \to \infty$ (we'll talk more about this when we deal with
  asymptotes).

  {\em Performance review}: $17$ out of $24$ got this. $4$ chose (D),
  $3$ chose (E).

  {\em Historical note 1}: $7$ out of $20$ students got this
  correct. $6$ chose (D), $5$ chose (C), $2$ chose (A).

  {\em Historical note 2}: In the last appearance in a 153 quiz, $16$
  out of $28$ people got this correct. $7$ people chose (A), $2$
  people each chose (C) and (E), and $1$ person chose (D).
\item For which of the following specifications is there {\bf no
  continuous function} satisfying the specifications? {\em Earlier
  scores: $7/14$, $21/28$}

  \begin{enumerate}[(A)]
  \item Domain $(0,1)$ and range $(0,1)$
  \item Domain $[0,1]$ and range $(0,1)$
  \item Domain $(0,1)$ and range $[0,1]$
  \item Domain $[0,1]$ and range $[0,1]$
  \item None of the above, i.e., we can get a continuous function for
    each of the specifications.
  \end{enumerate}

  {\em Answer}: Option (B)

  {\em Explanation}: By the extreme value theorem, any continuous
  function on a closed bounded interval must attain its maximum and
  minimum, and hence its image cannot be an open interval.

  {\em The other choices}:

  For options (A) and (D), we can pick the identity functions $f(x) :=
  x$ on the respective domains.

  For option (C), we can pick the function $f(x) := \sin^2(2\pi x)$ on
  the domain $(0,1)$. 

  {\em Performance review}: $20$ out of $24$ got this. $2$ chose (E),
  $1$ each chose (A) and (C).

  {\em Historical note 1}: $7$ out of $20$ people got this
  correct. $7$ chose (E), $4$ chose (C), $1$ each chose (A) and (D).

  {\em Historical note 2}: In the last appearance in a 153 quiz, $21$
  out of $28$ people got this correct. $5$ people chose (C) and $2$
  people chose (E).

\item Suppose $f$ is a continuously differentiable function from the
  open interval $(0,1)$ to $\R$. Suppose, further, that there are
  exactly $14$ values of $c$ in $(0,1)$ for which $f(c) = 0$. What can
  we say is {\bf definitely true} about the number of values of $c$ in
  the open interval $(0,1)$ for which $f'(c) = 0$? {\em Earlier scores:
  $7/15$, $19/28$}

  \begin{enumerate}[(A)]
  \item It is at least $13$ and at most $15$.
  \item It is at least $13$, but we cannot put any upper bound on it
    based on the given information.
  \item It is at most $15$, but we cannot put any lower bound (other
  than the meaningless bound of $0$) based on the given information.
  \item It is at most $13$.
  \item It is at least $15$.
  \end{enumerate}

  {\em Answer}: Option (B)

  {\em Explanation}: Suppose the zeros of $f$ are $a_1 < a_2 < \dots <
  a_{14}$. By Rolle's theorem, there is {\em at least one} zero of
  $f'$ between each $a_i$ and $a_{i+1}$. There may be more, since
  Rolle's theorem gives only a {\em lower} bound. This gives thirteen
  solutions $c$ to $f'(c) = 0$.

  Note that in order to apply Rolle's theorem, it is enough to be
  given that $f$ is differentiable, so the additional hypothesis that
  $f'$ is continuous is not necessary.

  {\em Performance review}: $23$ out of $24$ got this. $1$ chose (A).

  {\em Historical note 1}: $12$ out of $20$ people got this
  correct. $4$ chose (A), $2$ chose (E), $1$ each chose (C) and (D).

  {\em Historical note 2}: In the last appearance in a 153 quiz, $19$
  out of $28$ people got this correct. $5$ people chose (A), $2$
  people chose (C), and $1$ person each chose (D) and (E).
\item Consider the function $f(x) := \lbrace\begin{array}{rl} x, & 0
  \le x \le 1/2 \\ x - (1/7), & 1/2 < x \le 1 \\\end{array}$. Define by
  $f^{[n]}$ the function obtained by iterating $f$ $n$ times, i.e.,
  the function $f \circ f \circ f \circ \dots \circ f$ where $f$
  occurs $n$ times. What is the smallest $n$ for which $f^{[n]} =
  f^{[n + 1]}$? {\em Earlier score: $3/16$}

  \begin{enumerate}[(A)]
  \item $1$
  \item $2$
  \item $3$
  \item $4$
  \item $5$
  \end{enumerate}

  {\em Answer}: Option (D)

  {\em Explanation}: We need to iterate $f$ enough times that
  everything gets inside $[0,1/2]$, after which it becomes
  stable. Note that each time, the value goes down by $1/7$. Thus, for
  any $x \le 1$, we need at most four steps to bring it in $[0,1/2]$,
  with the upper bound of $4$ being attained for $1$.

  {\em Performance review}: $14$ out of $24$ got this. $5$ chose (C),
  $4$ chose (A), $1$ chose (E).

  {\em Historical note 1}: $4$ out of $20$ people got this
  correct. $7$ chose (B), $7$ chose (C), $1$ each chose (A) and (E).

  {\em Historical note 2}: In the last appearance in a 153 quiz, $10$
  out of $28$ people got this correct. $9$ people chose (C), $5$
  people chose (B), $2$ people chose (A), and $2$ people left the
  question blank.
\item Suppose $f$ and $g$ are functions $(0,1)$ to $(0,1)$ that are
  both right continuous on $(0,1)$. Which of the following is {\em
  not} guaranteed to be right continuous on $(0,1)$? {\em Earlier
  scores: $3/11$, $9/14$, $20/28$}

  \begin{enumerate}[(A)]
  \item $f + g$, i.e., the function $x \mapsto f(x) + g(x)$
  \item $f - g$, i.e., the function $x \mapsto f(x) - g(x)$
  \item $f \cdot g$, i.e., the function $x \mapsto f(x)g(x)$
  \item $f \circ g$, i.e., the function $x \mapsto f(g(x))$
  \item None of the above, i.e., they are all guaranteed to be right
    continuous functions
  \end{enumerate}

  {\em Answer}: Option (D)

  {\em Explanation}: See the explanation for Question 2 on the October
  1 quiz. Note that that quiz uses left continuity, but the example
  can be adapted to right continuity.

  {\em Performance review}: $18$ out of $24$ got this. $3$ chose (B),
  $2$ chose (E), $1$ chose (C).

  {\em Historical note 1}: $7$ out of $20$ people got this
  correct. $6$ chose (E), $5$ chose (C), $2$ chose (B).

  {\em Historical note 2}: In the last appearance in a 153 quiz, $20$ out
  of $28$ people got this correct. $4$ people chose (C), $2$ people
  chose (B), $1$ person chose (E), and $1$ person left the question
  blank.
\item Suppose $f$ and $g$ are increasing functions from $\R$ to
  $\R$. Which of the following functions is {\em not} guaranteed to be
  an increasing functions from $\R$ to $\R$? {\em Earlier scores:
  $1/15$, $9/16$, $18/28$}

  \begin{enumerate}[(A)]

  \item $f + g$
  \item $f \cdot g$
  \item $f \circ g$
  \item All of the above, i.e., none of them is guaranteed to be increasing.
  \item None of the above, i.e., they are all guaranteed to be increasing.
  \end{enumerate}

  {\em Answer}: Option (B)

  {\em Explanation}: The problem with option (B) arises when one or
  both functions take negative values. For instance, consider the case
  $f(x) := x$ and $g(x) := x$. Both are increasing functions on all of
  $\R$. However, the pointwise product is the function $x \mapsto
  x^2$, which is a decreasing function for negative $x$.

  Formally, the issue is that we cannot multiply inequalities of the
  form $A < B$ and $C < D$ unless we are guaranteed to be working with
  positive numbers.

  {\em The other choices}:

  Option (A): For any $x_1 <
  x_2$, we have $f(x_1) < f(x_2)$ and $g(x_1) < g(x_2)$. Adding up, we
  get $f(x_1) + g(x_1) < f(x_2) + g(x_2)$, so $(f + g)(x_1) < (f + g)(x_2)$.

  Option (C): For any $x_1 < x_2$, we have $g(x_1) < g(x_2)$ since $g$
  is increasing. Now, we use the factthat $f$ is increasing to compare
  its values at the two points $g(x_1)$ and $g(x_2)$, and we get
  $f(g(x_1)) < f(g(x_2))$. We thus get $(f \circ g)(x_1) < (f \circ
  g)(x_2)$.

  {\em Performance review}: $17$ out of $24$ got this. $4$ chose (E),
  $1$ each chose (A), (C), and (D).

  {\em Historical note 1}: $8$ out of $20$ people got this
  correct. $7$ chose (E), $4$ chose (C), $1$ chose (D).

  {\em Historical note 2}: In the last appearance in a 153 quiz, $18$
  out of $28$ people got this correct. $6$ people chose (E) and $4$
  people chose (C).

\item Suppose $F$ and $G$ are two functions defined on $\R$ and $k$ is
  a natural number such that the $k^{th}$ derivatives of $F$ and $G$
  exist and are equal on all of $\R$. Then, $F - G$ must be a
  polynomial function. What is the {\bf maximum possible degree} of $F
  - G$?  (Note: Assume constant polynomials to have degree zero) {\em
  Earlier score: $6/16$, $10/28$}

  \begin{enumerate}[(A)]
  \item $k - 2$
  \item $k - 1$
  \item $k$
  \item $k + 1$
  \item There is no bound in terms of $k$.
  \end{enumerate}

  {\em Answer}: Option (B)

  {\em Explanation}: $F$ and $G$ having the same $k^{th}$ derivative
  is equivalent to requiring that $F - G$ have $k^{th}$ derivative
  equal to zero. For $k = 1$, this gives constant functions
  (polynomials of degree $0$). Each time we increment $k$, the degree
  of the polynomial could potentially go up by $1$. Thus, the answer
  is $k - 1$.

  {\em Performance review}: $21$ out of $24$ got this. $2$ chose (A),
  $1$ chose (E).

  {\em Historical note 1}: $8$ out of $20$ people got this
  correct. $6$ chose (D), $3$ chose (C), $2$ chose (E), $1$ chose (A).

  {\em Historical note 2}: In the last appearance in a 153 quiz, $10$
  out of $28$ people got this correct. $5$ people each chose (D) and
  (E) and $4$ people each chose (A) and (C).

  {\em Action point}: This is a question you really {\em should} get
  correct!
\item Suppose $f$ is a continuous function on $\R$. Clearly, $f$ has
  antiderivatives on $\R$. For all but one of the following
  conditions, it is possible to guarantee, without any further
  information about $f$, that there exists an antiderivative $F$
  satisfying that condition. {\bf Identify the exceptional condition}
  (i.e., the condition that it may not always be possible to
  satisfy). {\em Earlier score: $3/16$, $10/28$}

  \begin{enumerate}[(A)]
  \item $F(1) = F(0)$.
  \item $F(1) + F(0) = 0$.
  \item $F(1) + F(0) = 1$.
  \item $F(1) = 2F(0)$.
  \item $F(1)F(0) = 0$.
  \end{enumerate}

  {\em Answer}: Option (A)

  {\em Explanation}: Suppose $G$ is an antiderivative for $f$. The
  general expression for an antiderivative is $G + C$, where $C$ is
  constant. We see that for options (b), (c), and (d), it is always
  possible to solve the equation we obtain to get one or more real
  values of $C$. However, (a) simplifies to $G(1) + C = G(0) + C$,
  whereby $C$ is canceled, and we are left with the statement $G(1) =
  G(0)$. If this statement is true, then {\em all} choices of $C$
  work, and if it is false, then {\em none} works. Since we cannot
  guarantee the truth of the statement, (a) is the exceptional
  condition.

  Another way of thinking about this is that $F(1) - F(0) = \int_0^1
  f(x) \, dx$, regardless of the choice of $F$. If this integral is
  $0$, then any antiderivative works. If it is not zero, no
  antiderivative works.

  {\em Performance review}: $18$ out of $24$ got this. $3$ chose (C),
  $2$ chose (E), $1$ chose (B).

  {\em Historical note 1}: $2$ out of $20$ people got this
  correct. $10$ chose (E), $4$ chose (D), $3$ chose (C), $1$ chose
  (B).
  
  {\em Historical note 2}: In the last appearance in a 153 quiz, $10$
  out of $28$ people got this correct. $6$ people chose (B), $5$
  people chose (E), $4$ people chose (D), $2$ people chose (C), and
  $1$ person left the question blank.
\item Suppose $F$ is a function defined on $\R \setminus \{ 0 \}$ such
  that $F'(x) = -1/x^2$ for all $x \in \R \setminus \{ 0 \}$. Which of
  the following pieces of information is/are {\bf sufficient} to determine
  $F$ completely? Please see Options (D) and (E) before answering.
  \begin{enumerate}[(A)]
  \item The value of $F$ at any two positive numbers.
  \item The value of $F$ at any two negative numbers.
  \item The value of $F$ at a positive number and a negative number.
  \item Any of the above pieces of information is sufficient, i.e., we
    need to know the value of $F$ at any two numbers.
  \item None of the above pieces of information is sufficient.
  \end{enumerate}

  {\em Answer}: Option (C)

  {\em Explanation}: There are two open intervals: $(-\infty,0)$ and
  $(0,\infty)$, on which we can look at $F$. On each of these
  intervals, $F(x) = 1/x + $ a constant, but the constant for
  $(-\infty,0)$ may differ from the constant for $(0,\infty)$. Thus,
  we need the initial value information at one positive number and one
  negative number.

  {\em Performance review}: $16$ out of $24$ got this. $8$ chose (D).

  {\em Historical note 1}: $5$ out of $20$ people got this
  correct. $12$ chose (D), $3$ chose (A).

  {\em Historical note 2}: In the last appearance in a 153 quiz, $15$
  out of $28$ people got this correct. $9$ people chose (D), $2$
  people chose (E), and $1$ person each chose (A) and (B).

\item Suppose $F$ and $G$ are continuously differentiable functions on
  all of $\R$ (i.e., both $F'$ and $G'$ are continuous). Which of the
  following is {\bf not necessarily true}? {\em Earlier scores: $0$,
  $10/16$}

  \begin{enumerate}[(A)]
  \item If $F'(x) = G'(x)$ for all integers $x$, then $F - G$ is a
    constant function when restricted to integers, i.e., it takes the
    same value at all integers.
  \item If $F'(x) = G'(x)$ for all numbers $x$ that are not integers,
    then $F - G$ is a constant function when restricted to the set of
    numbers $x$ that are not integers.
  \item If $F'(x) = G'(x)$ for all rational numbers $x$, then $F - G$
    is a constant function when restricted to the set of rational
    numbers.
  \item If $F'(x) = G'(x)$ for all irrational numbers $x$, then $F -
    G$ is a constant function when restricted to the set of irrational
    numbers.
  \item None of the above, i.e., they are all necessarily true.
  \end{enumerate}

  {\em Answer}: Option (A).

  {\em Explanation}: The fact that the derivatives of two functions
  agree at integers says nothing about how the derivatives behave
  elsewhere -- they could differ quite a bit at other places. Hence,
  (A) is not necessarily true, and hence must be the right option. All
  the other options are correct as statements and hence cannot be the
  right option. This is because in all of them, the set of points
  where the derivatives agree is {\em dense} -- it intersects every
  open interval. So, continuity forces the functions $F'$ and $G'$ to
  be equal everywhere, forcing $F - G$ to be constant everywhere.

  {\em Performance review}: $15$ out of $24$ got this. $6$ chose (E),
  $3$ chose (B).

  {\em Historical note 1}: $7$ out of $20$ people got this
  correct. $7$ chose (E), $3$ chose (D), $3$ chose (C).

  {\em Historical note 2}: In the last appearance in a 153 quiz, $11$
  out of $28$ people got this correct. $8$ people chose (E), $5$
  people chose (C), and $2$ people each chose (B) and (D).

\end{enumerate}

\end{document}