\documentclass[10pt]{amsart}

%Packages in use
\usepackage{fullpage, hyperref, vipul, enumerate}

%Title details
\title{Take-home class quiz solutions: due Wednesday February 6: Multivariable function basics continued}
\author{Math 195, Section 59 (Vipul Naik)}
%List of new commands

\begin{document}
\maketitle

\section{Performance review}

$26$ people took this quiz. The score distribution was as follows:

\begin{itemize}
\item Score of $0$: $4$ people
\item Score of $1$: $6$ people
\item Score of $2$: $9$ people
\item Score of $3$: $5$ people
\item Score of $4$: $2$ people
\end{itemize}

The question-wise answers and performance review were as follows:

\begin{enumerate}
\item Option (C): $12$ people 
\item Option (B): $13$ people
\item Option (E): $15$ people
\item Option (B): $7$ people
\end{enumerate}

\section{Solutions}
\begin{enumerate}
\item Suppose $F$ is an additively separable function of two variables
  $x$ and $y$ that is defined everywhere, i.e., there exist functions
  $f$ and $g$ of one variable, both defined on all of $\R$, such that
  $F(x,y) = f(x) + g(y)$ for all $x,y \in \R$.
 
  We call two curves {\em parallel} if there is a vector by which we
  can translate all the points in one curve to get precisely the other
  curve.

  Consider the following three statements:

  (i) All curves obtained as the intersections of the graph of $F$
  with planes parallel to the $xy$-plane are parallel to each other.

  (ii) All curves obtained as the intersections of the graph of $F$
  with planes parallel to the $xz$-plane are parallel to each other.

  (iii) All curves obtained as the intersections of the graph of $F$
  with planes parallel to the $yz$-plane are parallel to each other.

  Which of the statements (i)-(iii) is/are necessarily true?

  \begin{enumerate}[(A)]
  \item All of (i), (ii), and (iii) are true.
  \item Both (i) and (ii) are true but (iii) need not be true.
  \item Both (ii) and (iii) are true but (i) need not be true.
  \item Both (i) and (iii) are true but (ii) need not be true.
  \item (i) is true but (ii) and (iii) need not be true.
  \end{enumerate}

  {\em Answer}: Option (C)

  {\em Explanation}: 

  (i): The intersections are level curves, but these need not be
  parallel to each other. In fact, they could be different sizes, such
  as with $x^2 + y^2$.

  (ii): The intersection with a plane of the form $y = y_0$ gives the
  graph of a function $x \mapsto F(x,y_0) = f(x) + g(y_0)$. Note that
  all the functions whose graphs are obtained by such restrictions
  just look like the graph of $f$, translated to different $z$-heights
  and $y$-locations. Thus, they are parallel to one another. Note that
  we use additive separability in this reasoning.

  (iii): Similar reasoning as with (ii).

  {\em Performance review}: $12$ out of $26$ got this. $7$ chose (A),
  $6$ chose (E), $1$ chose (B).

\item Suppose $f$ is a continuous function of two variables $x$ and
  $y$, defined on the entire $xy$-plane. Suppose further that $f$ is
  increasing in $x$ for each fixed value of $y$, and that $f$ is
  increasing in $y$ for every fixed value of $x$. Which of the
  following is the most plausible description of the level curves of
  $f$ in the $xy$-plane? {\em Note: You might wish to take an
  extremely simple example, e.g., an additively separable function
  where each of the pieces is the simplest possible increasing
  function you can think of}.

  \begin{enumerate}[(A)]
  \item They are all upward-sloping, i.e., they are of the form $y =
    g(x)$ with $g$ an increasing function.
    \item They are all downward-sloping, i.e., they are of the form $y
      = g(x)$ with $g$ a decreasing function.
    \item They look like closed loops (e.g., circles).
    \item They look like graphs of functions with a unique local and
      absolute minimum (such as the parabola $y = x^2$, though the
      actual picture may be different).
    \item They look like graphs of functions with a unique local and
      absolute maximum (such as the parabola $y = -x^2$, though the
      actual function may be different).
  \end{enumerate}

  {\em Answer}: Option (B)

  {\em Explanation}: Since $f$ is increasing in $x$ and in $y$, it
  means that an increase in the $x$-value must be compensated by a
  decrease in the $y$-value to keep the output constant.

  The example $f(x,y) := x + y$ is illustrative. The level curves of
  these are downward-sloping straight lines.

  {\em Performance review}: $13$ out of $26$ got this. $7$ chose (A),
  $3$ each chose (C) and (D).
\item What do the level curves of the function $f(x,y) := \sin(x + y)$
  look like for output value in $[-1,1]$? Note that all these level
  curves are being considered as curves in the $xy$-plane. {\em Note:
  This builds upon the idea of Question 3 of the previous quiz.}

  \begin{enumerate}[(A)]
  \item Each level curve is a single line.
  \item Each level curve is a union of two intersecting lines.
  \item Each level curve is a union of two distinct parallel lines.
  \item Each level curve is a union of infinitely many concurrent
    lines (i.e., infinitely many lines, all passing through the same
    point).
  \item Each level curve is a union of infinitely many distinct
    parallel lines (i.e., infinitely many lines, all parallel to each
    other).
  \end{enumerate}

  {\em Answer}: Option (E)

  {\em Explanation}: For $C \in [-1,1]$, we need to solve $\sin(x + y)
  = C$. We first find all solutions $u$ to $\sin u = C$, which is a
  countably infinite subset of $\R$. Then, for each such $u$, we get
  the line $x + y = u$ in the $xy$-plane. All these lines are parallel
  to each other, and have slope $-1$.

  {\em Performance review}: $15$ out of $26$ got this. $6$ chose (A),
  $4$ chose (D), $1$ chose (B).

\item Suppose $f$ and $g$ are both continuous functions of two
  variables $x$ and $y$, both defined on all of $\R^2$, and such that
  $f(x,y) + g(x,y)$ is a constant $C$. What is the relation between
  the level curves of $f$ and the level curves of $g$, all drawn in
  the $xy$-plane?

  \begin{enumerate}[(A)]
  \item Every level curve of $f$ is a level curve of $g$ and vice
    versa, with the same level value for both functions.
  \item Every level curve of $f$ is a level curve of $g$ and vice
    versa, but the value for which it is a level curve may be different
    for the two functions.
  \item The level curves of $f$ need not be precisely the same as the
    level curves of $g$, but we can go from one set of level curves to
    the other via a parallel translation.
  \item Each level curve of $f$ can be obtained by reflecting a
    suitable level curve of $g$ about a suitable line in the
    $xy$-plane.
  \item Each level curve of $f$ can be obtained by reflecting a
    suitable level curves of $g$ about a suitable line in the
    $xy$-plane and then performing a suitable translation.
  \end{enumerate}

  {\em Answer}: Option (B)

  {\em Explanation}: Any level curve of the form $f(x,y) = k$
  coincides with the level curve $g(x,y) = C - k$. Note that unless $k
  = C/2$, the level values for the two curves are different.

  {\em Performance review}: $7$ out of $26$ got this. $8$ chose (D),
  $6$ chose (C), $3$ chose (A), $2$ chose (E).
\end{enumerate}

\end{document}