\documentclass[10pt]{amsart}

%Packages in use
\usepackage{fullpage, hyperref, vipul, enumerate}

%Title details
\title{Diagnostic in-class quiz: due Monday November 25: Subspace, basis, and dimension}
\author{Math 196, Section 57 (Vipul Naik)}
%List of new commands

\begin{document}
\maketitle

Your name (print clearly in capital letters): $\underline{\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad}$

{\bf PLEASE DO NOT DISCUSS {\em ANY} QUESTIONS.}

This quiz covers material related to the {\tt Linear dependence, bases
  and subspaces} notes corresponding to Sections 3.2 and 3.3 of the
text.

Keep in mind the following facts. Suppose $T:\R^m \to \R^n$ is a
linear transformation. Suppose $A$ is the matrix for $T$, so that
$T(\vec{x}) = A\vec{x}$ for all $\vec{x} \in \R^m$. Then, $A$ is a $n
\times m$ matrix. Further, the following are true:

\begin{itemize}
\item The dimension of the image of $T$ equals the rank of $A$.
\item The dimension of the kernel of $T$, called the {\em nullity} of
  $A$, is $m$ minus the rank of $A$.
\end{itemize}

\begin{enumerate}
\item {\em Do not discuss this!}: Suppose $T:\R^m \to \R^n$ is a
  linear transformation. What is the best we can say about the
  dimension of the image of $T$?

  \begin{enumerate}[(A)]
  \item It is at least $0$ and at most $\min \{m,n\}$. However, we
    cannot be more specific based on the given information.
  \item It is at least $0$ and at most $\max \{m,n \}$. However, we
    cannot be more specific based on the given information.
  \item It is at least $\min \{ m,n \}$ and at most $\max \{ m,n
    \}$. However, we cannot be more specific based on the given information.
  \item It is at least $\min \{ m,n \}$ and at most $m + n$. However,
    we cannot be more specific based on the given information.
  \item It is at least $\max \{ m,n \}$ and at most $m + n$. However,
    we cannot be more specific based on the given information.
  \end{enumerate}

  \vspace{0.1in}
  Your answer: $\underline{\qquad\qquad\qquad\qquad\qquad\qquad\qquad}$

\item {\em Do not discuss this!}: Suppose $T_1,T_2:\R^m \to \R^n$ are
  linear transformations. Suppose the images of $T_1$ and $T_2$ have
  dimensions $d_1$ and $d_2$ respectively. What can we say about the
  dimension of the image of $T_1 + T_2$? Assume that both $m$ and $n$
  are larger than $d_1 + d_2$.

  \begin{enumerate}[(A)]
  \item It is precisely $|d_2 - d_1|$.
  \item It is precisely $\min \{ d_1, d_2 \}$.
  \item It is precisely $\max \{ d_1, d_2 \}$.
  \item It is precisely $d_1 + d_2$.
  \item Based on the information, it could be any integer $r$ with
    $|d_2 - d_1| \le r \le d_1 + d_2$.
  \end{enumerate}

  \vspace{0.1in}
  Your answer: $\underline{\qquad\qquad\qquad\qquad\qquad\qquad\qquad}$

\item {\em Do not discuss this!}: Suppose $V_1$ and $V_2$ are
  subspaces of $\R^n$. We define the sum $V_1 + V_2$ as the subset of
  $\R^n$ comprising all vectors that can be expressed as a sum of a
  vector in $V_1$ and a vector in $V_2$. Define $V_1 \cup V_2$ as the
  set-theoretic union of $V_1$ and $V_2$, i.e., the set of all vectors
  that are either in $V_1$ or in $V_2$. What can we say about these?

  \begin{enumerate}[(A)]
  \item $V_1 \cup V_2 = V_1 + V_2$ and it is a subspace of $\R^n$.
  \item $V_1 \cup V_2$ is contained in $V_1 + V_2$ and both are subspaces of $\R^n$.
  \item $V_1 \cup V_2$ is contained in $V_1 + V_2$, and $V_1 + V_2$ is
    a subspace of $\R^n$. $V_1 \cup V_2$ is generally not a subspace
    of $\R^n$ (though it might be in special cases).
  \item $V_1 \cup V_2$ contains $V_1 + V_2$, and both are subspaces of $\R^n$.
  \item $V_1 \cup V_2$ contains $V_1 + V_2$, and $V_1 \cup V_2$ is a
    subspace of $\R^n$. $V_1 + V_2$ is generally not a subspace of
    $\R^n$ (though it might be in special cases).
  \end{enumerate}

  Your answer: $\underline{\qquad\qquad\qquad\qquad\qquad\qquad\qquad}$

\end{enumerate}
\end{document}
