\documentclass[10pt]{amsart}
\usepackage{fullpage,hyperref,vipul,graphicx}
\title{Geometry of linear transformations}
\author{Math 196, Section 57 (Vipul Naik)}

\begin{document}
\maketitle

{\bf Corresponding material in the book}: Section 2.2.

\section*{Executive summary}

\begin{enumerate}
\item There is a concept of {\em isomorphism} as something that
  preserves essential structure or feature, where the concept of
  isomorphism depends on what feature is being preserved.
\item There is a concept of {\em automorphism} as an isomorphism from
  a structure to itself. We can think of automrohpisms of a structure
  as {\em symmetries} of that structure.
\item Linear transformations have already been defined. An {\em affine
  linear transformation} is something that preserves lines and ratios
  of lengths within lines. Any affine linear transformation is of the
  form $\vec{x} \mapsto A\vec{x} + \vec{b}$. For the transformation to
  be linear, we need $\vec{b}$ to be the zero vector, i.e., the
  transformation must send the origin to the origin. If $A$ is the
  identity matrix, then the affine linear transformation is termed a
  {\em translation}.
\item A linear {\em isomorphism} is an invertible linear
  transformation. For a linear isomorphism to exist from $\R^m$ to
  $\R^n$, we must have $m = n$. An affine linear isomorphism is an
  invertible affine linear transformation.
\item A linear automorphism is a linear isomorphism from $\R^n$ to
  itself. An affine linear automorphism is an affine linear
  isomorphism from $\R^n$ to itself.
\item A self-isometry of $\R^n$ is an invertible function from $\R^n$
  to itself that preserves Euclidean distance. Any self-isometry of
  $\R^n$ must be an affine linear automorphism of $\R^n$.
\item A self-homothety of $\R^n$ is an invertible function from $\R^n$
  to itself that scales all Euclidean distances by a factor of
  $\lambda$, where $\lambda$ is the factor of homothety. We can think
  of self-isometries precisely as the self-homotheties by a factor of
  $1$. Any self-homothety of $\R^n$ must be an affine linear
  automorphism of $\R^n$.
\item Each of these forms a group: the affine linear automorphisms of
  $\R^n$, the linear automorphisms of $\R^n$, the self-isometries of
  $\R^n$, the self-homotheties of $\R^n$.
\item For a linear transformation, we can consider something called
  the {\em determinant}. For a $2 \times 2$ linear transformation with
  matrix

  $$\left[\begin{matrix} a & b \\ c & d \\\end{matrix}\right]$$

  the determinant is $ad - bc$.

  We can also consider the {\em trace}, defined as $a + d$ (the sum of
  the diagonal entries).
\item The trace generalizes to $n \times n$ matrices: it is the sum of
  the diagonal entries. The determinant also generalizes, but the
  formula becomes more complicated.
\item The determinant for an affine linear automorphism can be defined
  as the determinant for its linear part (the matrix).
\item The sign of the determinant being positive means the
  transformation is orientation-preserving. The sign of the
  determinant being negative means the transformation is
  orientation-reversing.
\item The magnitude of the determinant gives the factor by which
  volumes are scaled. In the case $n = 2$, it is the factor by which
  areas are scaled.
\item The determinant of a self-homothety with factor of homothety
  $\lambda$ is $\pm \lambda^n$, with the sign depending on whether it
  is orientation-preserving or orientation-reversing.
\item Any self-isometry is volume-preserving, so it has determinant
  $\pm 1$, with the sign depending on whether it is
  orientation-preserving or orientation-reversing.
\item For $n = 2$, the orientation-preserving self-isometries are
  precisely the translations and rotations. The ones fixing the origin
  are precisely rotations centered at the origin. These form groups.
\item For $n = 2$, the orientation-reversing self-isometries are
  precisely the reflections and glide reflections. The ones fixing the
  origin are precisely reflections about lines passing through the
  origin.
\item For $n = 3$, the orientation-preserving self-isometries fixing
  the origin are precisely the rotations about axes through the
  origin. The overall classification is more complicated.
\end{enumerate}

\section{Geometry of linear transformations}

\subsection{Geometry is secondary, but helps build intuition}

Our focus on linear transformations so far has been {\em
  information-centric}: they help reframe existing pieces of
information in new ways, allowing us to extract things that are
valuable to us. This information-centric approach undergirds the
applicability of linear algebra in the social sciences.

We will now turn to a geometric perspective on linear transformations,
restricting largely to the case $n = 2$ for many aspects of our
discussion. A lot of this geometry is more parochial than the
information-centric approach, and is not too important for the
application of linear algebra to the social sciences. The main reason
the geometry is valuable to us is that it can help build intuition
regarding what is going on with the algebraic operations and hence
offer another way of ``sanity checking'' our algebraic intuitions. As
a general rule, always try to come up with reasons that are algebraic
and information-centric, but in cases where we can perform sanity
checks using geometric intuitions, use them.

\subsection{Composition of transformations}

If $f:A \to B$ is a function and $g:B \to C$ is a function, we can
make sense of the composite $g \circ f$. The key feature needed to
make sense of the composite is that the co-domain (the target space)
of the function applied first (which we write on the right) must equal
the domain (the starting space) of the function applied next (which we
write later).

This means that if we have a collection of maps all from a space to
{\em itself}, it makes sense to compose any two maps in the
collection. We can even compose more than two maps if necessary.

The geometry of linear transformations that we discuss here is in the
context of transformations from $\R^n$ to $\R^n$. Here, we can
compose, and if the transformations are bijective, also invert.

A small note is important here. We often see maps from $\R^n$ to
$\R^n$ where, even though both the domain and co-domain space have the
same dimension, they do not represent the ``same space''
conceptually. For instance, one side may be measuring masses, while
the other side may be measuring prices. In this case, composing
multiple such maps does not make sense because even though the domain
and co-domain are mathematically $\R^n$, they are conceptually
different.

\section{Particular kinds of transformations in linear algebra}

\subsection{The concept of isomorphism and automorphism}

We will briefly describe two central mathematical concepts called {\em
  isomorphism} and {\em automorphism}. You are not expected to
understand these concepts, but they help demystify some of the
following discussion.

Understanding the concept of {\em isomorphism} is central to
abstraction and to human intelligence, even though the word is not too
well-known outside of mathematics and philosophy. When looking at
different structures, we may be interested in whether they have {\em a
  certain feature in common}. For instance, when looking at sets, we
may be interested in judging them by the number of elements in
them. If we care only about size, then in our tunnel vision, all other
aspects of the set are irrelevant. For our purposes, then, a set of
three rabbits is effectively the same as a set of three lions, or a
set of three stars. As another related example, suppose the only thing
you care about potential mates in the dating market is their bank
balance. In that case, two potential mates with the same bank balance
are effectively the same, i.e., they are isomorphic.

This kind of abstraction is crucial to humans being able to understand
numbers in the first place. If you crossed that hurdle back at the age
of $3$, $4$, $5$, or $6$ (or whenever you understood the idea of
counting), it's high time you went a step further.

One way of capturing the idea that two sets have the same size is as
follows. We say that two sets have the same size if it is possible to
construct a bijective function from one set to the other. In fact,
roughly speaking, this is the {\em only} way to define the concept of
``same size''. In other words, we can {\em define} the size (technical
term: cardinality) of a set as that attribute that is preserved by
bijections and is different for sets if there is no bijection between
them.

Another way of framing this is to christen bijective functions as {\em
  isomorphisms of sets}. In other words, a function from a set to a
(possibly same, possibly different) set is termed an isomorphism of
sets if it is a bijective function of sets. ``Iso+morph'' stands for
``same shape'' and signifies that the bijective function preserves the
shape.

Why is this important? Instead of caring only about size, we can
ratchet up our caring levels to care about more structural aspects of
the sets we are dealing with. If we do so, our definition of
``isomorphism'' will become correspondingly more stringent, since it
will require preserving more of the structure.

A related notion to isomorphism is that of {\em automorphism}. An
automorphism is an isomorphism (in whatever sense the term is being
used) from a set to itself.

Whatever our definition of isomorphism and automorphism, the identity
map from a set to itself is always an automorphism.

Non-identity automorphisms from a set to itself signify {\em
  symmetries} of the set. This will become clearer as we proceed.

\subsection{Linear isomorphisms and automorphisms}

A {\em linear isomorphism} is defined as a bijective linear
transformation. We can think of a linear isomorphism as a map that
preserves precisely all the ``linear structure'' of the set.

Note that for a linear transformation to be bijective, the dimensions
of the start and end space are the same. Explicitly, if $T:\R^m \to
\R^n$ is a bijective linear transformation, then $m = n$. Also, the
matrix for $T$ has full rank. It is also a square matrix, since $m =
n$.

In other words, the dimension of a vector space is invariant under
linear isomorphisms. In fact, the relationship between linear
isomorphisms and dimension is similar to the relationship between set
isomorphisms and cardinality (set size).

A {\em linear automorphism} is defined as a linear isomorphism from a
vector space $\R^n$ to itself. Based on the above discussion, you
might believe that every linear isomorphism must be a linear
automorphism. This is not quite true. The main caveat at the moment
(there will be more later) is that $\R^n$ and $\R^n$ could refer to
different spaces depending on what kinds of things we are storing
using the real numbers (for instance, one $\R^n$ might be masses, the
other $\R^n$ might be prices, and so on). A linear isomorphism between
a $\R^n$ of one sort and a $\R^n$ of another sort should not rightly
be considered a linear automorphism.

\subsection{Affine isomorphisms}

An {\em affine linear transformation} is a function that preserves
collinearity and ratios within lines, $T:\R^m \to \R^n$ is affine if
the image of any line (not necessarily through the origin) in $\R^m$
is a line (not necessarily through the origin) in $\R^n$, and
moreover, it preserves the ratios of lengths within each line. So in
particular, if $\vec{x},\vec{y} \in \R^m$, then $T(a\vec{x} + (1 -
a)\vec{y}) = aT(\vec{x}) + (1 - a)T(\vec{y})$. In particular, it
preserves midpoints.

Note that any linear transformation is affine linear. The main thing
we get by allowing ``affine'' is that the origin need not go to the
origin. An important class of affine linear transformations that are
not linear is the class of {\em translations}. Explicitly, for a
nonzero vector $\vec{v} \in \R^n$, the function $T:\R^n \to \R^n$
given by $\vec{x} \mapsto \vec{x} + \vec{v}$ is affine linear but not
linear. In fact, all affine linear transformations can be obtained by
composing translations with linear transformations.

An {\em affine linear isomorphism} (or {\em affine isomorphism} for
short) is a bijective affine linear transformation.

An {\em affine linear automorphism} is an affine linear isomorphism
from a vector space to itself.

Every linear automorphism is an affine linear automorphism. Nonzero
translations give examples of affine linear automorphisms that are not
linear. In an important sense, these cover all the affine linear
automorphisms: every affine linear automorphism can be expressed as a
composite of a translation and a linear automorphism.

\subsection{Linear and affine linear: the special case of one dimension}

In the case of one dimension, a linear isomorphism is a function of
the form $x \mapsto mx, m \ne 0$. The matrix of this, viewed as a
linear transformation, is $[m]$. In other words, it is a linear
function with zero intercept.

An affine linear isomorphism is a function of the form $x \mapsto mx +
c$, $m \ne 0$. In other words, it is a linear function with the
intercept allowed to be nonzero.

\subsection{The general description of affine linear transformations}

A linear transformation $T:\R^m \to \R^n$ can be written in the form:

$$\vec{x} \mapsto A\vec{x} + \vec{b}$$

where $A$ is a $n \times m$ matrix and $\vec{v}$ is a vector in
$\R^n$. Explicitly, the matrix $A$ describes the linear transformation
part and the vector $\vec{b}$ describes the translation part. Notice
how this general description parallels and generalizes the description
in one dimension.

\subsection{Understanding how transformations and automorphisms behave intuitively}

In order to understand transformations and automorphisms using our
geometric intuitions, it helps to start off with some geometric
picture in the plane or Euclidean space that we are transforming, then
apply the transformation to it, and see what we get. It is preferable
to not take something {\em too} symmetric, because the less the
symmetry, the more easily we can discern what features the
transformation preserves and what features it destroys. Human stick
features with asymmetric faces may be a good starting point for
intuitive understanding, though more mundane figures like triangles
might also be reasonable.

\section{Euclidean geometry}

\subsection{Euclidean distance and self-isometries}

The geometry of $\R^n$ is largely determined by how we define distance
between points. The standard definition of distance is {\em Euclidean
  distance}. Explicitly, if $\vec{x}$ and $\vec{y}$ are in $\R^n$,
then the Euclidean distance between $\vec{x}$ and $\vec{y}$ is:

$$\sqrt{\sum_{i=1}^n (x_i - y_i)^2}$$

If a transformation preserves Euclidean distance, then it preserves
all the geometry that we care about. In particular, it preserves
shapes, sizes, angles, and other geometric features.

A bijective function from $\R^n$ to itself that preserves Euclidean
distance is termed a {\em self-isometry} of $\R^n$.

We now proceed to an explanation of why self-isometries of $\R^n$ must
necessarily be affine linear automorphisms of $\R^n$. The claim is
that if something preserves distance, it must preserve linearity.

One way of characterizing the fact that three points $A,B,C \in \R^n$
are collinear, with $B$ {\em between} $A$ and $C$, is that we get the
equality case for the triangle inequality. Explicitly:

$$\text{(the distance between $A$ and $B$)} + \text{(the distance between $B$ and $C$)} = \text{(the distance between $A$ and $C$)}$$

Suppose $T$ is a self-isometry of $\R^n$. Then, we have:

\begin{eqnarray*}
  \text{The distance between $T(A)$ and $T(B)$} & = & \text{The distance between $A$ and $B$}\\
  \text{The distance between $T(B)$ and $T(C)$} & = & \text{The distance between $B$ and $C$}\\
  \text{The distance between $T(A)$ and $T(C)$} & = & \text{The distance between $A$ and $C$}\\
\end{eqnarray*}

Combining all these, we get that:

$$\text{(the distance between $T(A)$ and $T(B)$)} + \text{(the distance between $T(B)$ and $T(C)$)} = \text{(the distance between $T(A)$ and $T(C)$)}$$

The conclusion is that $T(A)$, $T(B)$, and $T(C)$ are collinear with
$T(B)$ between $T(A)$ and $T(C)$. In other words, collinear triples of
points get mapped to collinear triples of points, so $T$ preserves
collinearity. Further, it obviously preserves ratios of lengths within
lines. Thus, $T$ is an affine linear automorphism of $\R^n$. The
upshot: every self-isometry is an affine linear automorphism.

Self-isometries preserve not just collinearity, but {\em all} the
geometric structure. What they are allowed to change is the location,
angling, and orientation. They do not affect the shape and size of
figures. In particular, they send triangles to congruent triangles.

\subsection{Self-homotheties}

A {\em self-homothety} or {\em similarity transformation} or {\em
  similitude transformation} is a bijective map $T:\R^n \to \R^n$ that
scales all distances by a fixed nonzero factor called the {\em ratio
  of similitude} or {\em factor of similitude} of $T$. For instance, a
self-homothety by a factor of $1/2$ will have the property that the
distance between $T(A)$ and $T(B)$ is half the distance between $A$
and $B$.

Self-isometries can be described as self-homotheties by a factor of
$1$.

Self-homotheties are affine linear automorphisms for roughly the same
reason that self-isometries are.

A special kind of self-homothety is a {\em dilation} about a point. A
dilation about the origin, for instance, would simply mean multiplying
all the position vectors of points by a fixed nonzero scalar. The
absolute value of that scalar will turn out to be the factor of
similitude. The matrix for such a dilation is a scalar matrix. For
instance, the matrix:

$$\left[\begin{matrix} 5 & 0 \\ 0 & 5 \\\end{matrix}\right]$$

represents a dilation by a factor of $5$ about the origin.

Self-homotheties send triangles to similar triangles.

\subsection{Other types of transformations}

One important type of linear transformation that is not a
self-homothety is a transformation with diagonal matrix where the
diagonal entries are not all the same. For instance, consider the
linear transformation with matrix the following diagonal matrix:

$$\left[\begin{matrix} 1 & 0 \\ 0 & 2 \\\end{matrix}\right]$$

This sends $\vec{e_1}$ to itself and sends $\vec{e_2}$ to
$2\vec{e_2}$. Pictorially, it keeps the $x$-axis as is and stretches
the $y$-axis by a factor of $2$. It distorts shapes, but preserves
linearity. It is not a self-homothety because of the different scaling
factors used for the axes.
\subsection{Group structure}

A collection of bijective functions from $\R^n$ to $\R^n$ is said to
form a {\em group} if it satisfies these three conditions:

\begin{itemize}
\item The composite of two functions in the collection is in the collection.
\item The identity function is in the collection.
\item The inverse to any function in the collection is in the collection.
\end{itemize}

The set of all automorphisms of any structure forms a group. Here is
the stylized argument:

\begin{itemize}
\item Automorphisms preserve some particular structural
  feature. Composing two automorphisms will therefore also preserve
  the structural feature.
\item The identity map preserves {\em everything}. Therefore, it must
  be an automorphism.
\item Since an automorphism preserves a specific structural feature,
  doing it backwards must also preserve the structural feature.
\end{itemize}

All the examples we have seen above give groups of linear
transformations. Explicitly:

\begin{itemize}
\item The set of all affine linear automorphisms of $\R^n$ is a group,
  because these are precisely the invertible functions that preserve
  the collinearity and ratios-within-lines structure.
\item The set of all linear automorphisms of $\R^n$ is a group,
  because these are precisely the invertible functions that preserve
  the linear structure.
\item The set of all self-isometries of $\R^n$ is a group, because
  these are precisely the invertible functions that preserve the
  Euclidean distance.
\item The set of all self-homotheties of $\R^n$ of a group, because
  these are precisely the invertible functions that preserve the
  ``ratios of Euclidean distances'' structure.
\end{itemize}

Further, these groups have containment relations:

Group of all self-isometries of $\R^n$ $\subseteq$ Group of all self-homotheties of $\R^n$ $\subseteq$ Group of all affine linear automorphisms of $\R^n$

And also, separately:

Group of all linear automorphisms of $\R^n$ $\subseteq$ Group of all affine linear automorphisms of $\R^n$

\section{The case of two dimensions}

If $n = 2$, we can obtain a relatively thorough understanding of the
various types of linear transformations. These are discussed in more
detail below.

\subsection{Trace and determinant}

There are two interesting invariants of $2 \times 2$ matrices, called
respectively the {\em trace} and {\em determinant}. The trace of a $2
\times 2$ matrix:

$$\left[\begin{matrix} a & b \\ c & d \\\end{matrix}\right]$$

is defined as the quantity $a + d$. It is the sum of the diagonal
entries. The significance of the trace is not clear right now, but
will become so later.

The other important invariant for $2 \times 2$ matrices is the {\em
  determinant}. The determinant of a $2 \times 2$ matrix:

$$\left[\begin{matrix} a & b \\ c & d \\\end{matrix}\right]$$

is defined as the quantity $ad - bc$. As we noted in a homework
exercise, the determinant is nonzero if and only if the matrix is
invertible.

Both the trace and the determinant generalize to $n \times n$
matrices. The trace of a $n \times n$ matrix is defined as the sum of
all the diagonal entries of the matrix. The determinant is defined in
a more complicated fashion.

The trace and determinant of a linear transformation are defined
respectively as the trace and determinant of the matrix for the linear
transformation.

For an affine linear transformation, the trace and determinant are
defines respectively as the trace and determinant of the linear part
of the transformation.

The role of the determinant is somewhat hard to describe, but it can
be split into two aspects:

\begin{itemize}
\item The absolute value of the determinant is the factor by which
  volumes multiply. In the case $n = 2$, it is the factor by which
  areas multiply. In particular, the determinant of a $2 \times 2$
  matrix is $\pm 1$ if and only if the corresponding linear
  transformation is area-preserving.
\item The sign of the determinant describes whether the linear
  transformation is orientation-preserving or orientation-reversing. A
  positive sign means orientation-preserving, whereas a negative sign
  means orientation-reversing. Here, {\em orientation-preserving}
  means that left-handed remains left-handed while right-handed
  remains right-handed. In contrast, {\em orientation-reversing} means
  that left-handed becomes right-handed while right-handed becomes
  left-handed. Note that any transformation that can be accomplished
  through rigid motions, i.e., through a continuous deformation of the
  identity transformation, must be orientation-preserving. The reason
  is that continuous change cannot suddenly change the orientation
  status.
\end{itemize}

\subsection{Justifying statements about the determinant using diagonal matrices}

Consider a linear transformation with diagonal matrix:

$$\left[\begin{matrix} a & 0 \\ 0 & d \\\end{matrix}\right]$$

The trace of this matrix is $a + d$ and the determinant is $ad$. Here
is the justification for both observations made above:

\begin{itemize}
\item The absolute value of the determinant is the factor by which
  volumes multiply: Think of a rectangle with sides parallel to the
  axes. The $x$-dimension gets multiplied by a factor of $|a|$ and the
  $y$-dimension gets multiplied by a factor of $|d|$. The area
  therefore gets multiplied by a factor of $|a||d|$ which is $|ad|$,
  the absolute value of the determinant.
\item The sign of the determinant describes whether the linear
  transformation is orientation-preserving or
  orientation-reversing. The sign of $a$ determines whether the
  $x$-direction gets flipped. The sign of $d$ determines whether the
  $y$-direction gets flipped. The sign of the product determines
  whether the overall orientation stays the same or gets reversed.
\end{itemize}

\subsection{Abstract considerations: determinants of self-homotheties}

Consider a linear transformation that is a self-homothety with factor
of similitude $\lambda$. This linear transformation scales all lengths
by a factor of $\lambda$. If $n = 2$ (i.e., we are in two dimensions)
then it scales all areas by a factor of $\lambda^2$. In particular:

\begin{itemize}
\item If it is orientation-preserving, then the determinant is $\lambda^2$.
\item If it is orientation-reversing, then the determinant is $-\lambda^2$.
\end{itemize}

In particular, for a self-{\em isometry}:

\begin{itemize}
\item If it is orientation-preserving, then the determinant is $1$.
\item If it is orientation-reversing, then the determinant is $-1$.
\end{itemize}

\subsection{Rotations}

A particular kind of transformation of interest in the two-dimensional
case is a {\em rotation}. A rotation is specified by two pieces of
information: a point (called the {\em center of rotation}) and an
angle (called the {\em angle of rotation}). The angle is defined only
up to additive multiples of $2\pi$, i.e., if two rotations have the
same center and their angles differ by a multiple of $2\pi$, then they
are actually the same rotation.

Note that we set a convention in advance that we will interpret
rotations in the counter-clockwise sense.

For a rotation whose angle of rotation is not zero (or more precisely,
is not a multiple of $2\pi$), the center of rotation is uniquely
determined by the rotation and is the only fixed point of the
rotation.

The rotation by an angle of $\pi$ about a point is termed a {\em half
  turn} about the point and can alternatively by thought of as {\em
  reflecting} relative to the {\em point}. This is not to be confused
with reflections about lines in $\R^2$.

All rotations are self-isometries of $\R^2$. Thus, they are affine
linear automorphisms. They are also area-preserving. They are also
orientation-preserving, since they can be obtained through continuous
rigid motions. However, unless the center of rotation is the origin,
the rotation is not a linear automorphism. Rotations centered at the
origin {\em are} linear transformations. We will now proceed to
describe rotations centered at the origin in matrix terms.

To describe a rotation centered at the origin, we need to describe the
images of the two standard basis vectors $\vec{e_1}$ and
$\vec{e_2}$. These images form the first and second column
respectively of the matrix describing the rotation as a linear
transformation.

Suppose the rotation is by an angle $\theta$. Then $\vec{e_1}$ goes to
the vector with coordinates $(\cos \theta,\sin \theta)$. $\vec{e_2}$
goes to the vector with coordinates $(-\sin \theta,\cos \theta)$. The matrix for the rotation is thus:

$$\left[\begin{matrix} \cos \theta & - \sin \theta \\ \sin \theta & \cos \theta \\\end{matrix}\right]$$

Note that the inverse to rotation about the origin by $\theta$ is
rotation by $-\theta$. Using the fact that $\cos$ is an even function
and $\sin$ is an odd function ,the matrix for the inverse operation
is:

$$\left[\begin{matrix} \cos \theta & \sin \theta \\ -\sin \theta & \cos \theta \\\end{matrix}\right]$$

Some specific rotations of interest are listed below:

\begin{tabular}{|l|l|}
  \hline
  Angle & Matrix of rotation\\\hline
  $0$ (no change) & $\left[\begin{matrix} 1 & 0 \\ 0 & 1 \\\end{matrix}\right]$\\\hline
  $\pi/4$ & $\left[\begin{matrix} 1/\sqrt{2} & -1/\sqrt{2} \\ 1/\sqrt{2} & 1/\sqrt{2} \\\end{matrix}\right]$\\\hline
  $\pi/2$ (right angle)& $\left[\begin{matrix} 0 & -1 \\ 1 & 0 \\\end{matrix}\right]$\\\hline
  $\pi$ (half turn) & $\left[\begin{matrix} -1 & 0 \\ 0 & -1 \\\end{matrix}\right]$\\\hline
  $3\pi/2$ (right angle clockwise) & $\left[\begin{matrix} 0 & 1 \\ -1 & 0 \\\end{matrix}\right]$ \\\hline
\end{tabular}

Let us try predicting the determinant of a rotation matrix
theoretically, then proceed to verify it computationally.

Theoretically, we know that rotations are both area-preserving (on
account of being self-isometries) and orientation-preserving (on
account of being realized through rigid motions). The area-preserving
nature tells us that the magnitude of the determinant is $1$, i.e.,
the determinant is $\pm 1$. The orientation-preserving nature tells us
that the determinant is positive. Combining these, we get that the
determinant must be $1$. Let's check this. The determinant of

$$\left[\begin{matrix} \cos \theta & - \sin \theta \\ \sin \theta & \cos \theta \\\end{matrix}\right]$$

is $\cos^2\theta + \sin^2\theta$, which is $1$.

\subsection{Reflections and orientation-reversal}

A reflection about a line does exactly what it is supposed to do: it
sends each point to another point such that the line of reflection is
the perpendicular bisector of the line segment joining them.

Every reflection is an affine linear automorphism. A reflection is a
linear automorphism if and only if the line of reflection passes
through the origin. If that is the case, we can write the matrix of
the linear transformation. Let's consider the case of a reflection
about the $x$-axis.

This reflection fixes all points on the $x$-axis, and sends all points
on the $y$-axis to their mirror images about the origin. In
particular, it sends $\vec{e_1}$ to $\vec{e_1}$ and sends $\vec{e_2}$
to $-\vec{e_2}$. 

The matrix of the linear transformation is:

$$\left[\begin{matrix} 1 & 0 \\ 0 & -1 \\\end{matrix}\right]$$

More generally, consider a reflection about a line through the origin
that makes an angle of $\theta/2$ counter-clockwise from the
$x$-axis. The reflection sends $\vec{e_1}$ to a vector making an angle
$\theta$ counter-clockwise from the horizontal, and sends $\vec{e_2}$
to the vector making an angle of $\theta - (\pi/2)$ counter-clockwise
from the horizontal. The matrix is thus:

$$\left[\begin{matrix} \cos \theta & \sin \theta \\ \sin \theta & - \cos \theta \\\end{matrix}\right]$$

This reflection matrix has trace zero. For the determinant, let us
first predict it theoretically, then verify it
computationally. Theoretically, we know that reflections are
self-isometries, hence they are area-preserving. So, the absolute
value of the determinant is $1$, and the determinant is $\pm
1$. Reflections are also orientation-reversing, so the determinant is
negative. Thus, the determinant must be $-1$. Let's check this. The
determinant of:

$$\left[\begin{matrix} \cos \theta & \sin \theta \\ \sin \theta & - \cos \theta \\\end{matrix}\right]$$

is $-\cos^2\theta - \sin^2\theta = -1$.
\subsection{Shear operations}

A shear operation is an operation where one axis is kept fixed, and
the other axis is ``sheared'' by having stuff from the fixed axis
added to it. 

For instance, suppose the $x$-axis is the fixed axis and
we add the standard basis vector for the $x$-axis to the
$y$-axis. Explicitly, $\vec{e_1}$ stays where it is, but $\vec{e_2}$
gets sent to $\vec{e_1} + \vec{e_2}$. The matrix of this looks like:

$$\left[\begin{matrix} 1 & 1 \\ 0 & 1 \\\end{matrix}\right]$$

Note that unlike translations, rotations, and reflections, shear
operations are not self-isometries.

More generally, we can have a shear of the form:

$$\left[\begin{matrix} 1 & \lambda \\ 0 & 1 \\\end{matrix}\right]$$

This sends $\vec{e_1}$ to $\vec{e_1}$ and sends $\vec{e_2}$ to
$\lambda\vec{e_1} + \vec{e_2}$.

We could also shear in the other direction:

$$\left[\begin{matrix} 1 & 0 \\ \lambda & 1 \\\end{matrix}\right]$$

Here, $\vec{e_2}$ is fixed, and $\vec{e_1}$ gets sends to $\vec{e_1} +
\lambda\vec{e_2}$.

The trace of a shear operation is $2$, and the determinant is
$1$. Thus, shear operations are both area-preserving and
orientation-preserving. This can be verified pictorially.

\subsection{Composites of various types, including glide reflections}

We have considered translations, rotations, and reflections. All of
these are self-isometries. Self-isometries form a group, so composing
things of these types should also give a self-isometry. Two
interesting questions:

\begin{itemize}
\item Is every self-isometry a translation, rotation, or reflection?
\item Can every self-isometry be obtained by composing translations,
  rotations, and reflections?
\end{itemize}

It turns out that the answers are respectively {\em no} and {\em
  yes}. Let's look at various sorts of composites:

\begin{enumerate}
\item {\em Composite of translations}: Translations form a subgroup of
  the group of all self-isometries of $\R^2$. In other words, the
  composite of two translations is a translation, the identity map is
  a translation (namely, by the zero vector) and the inverse of a
  translation is a translation.
\item {\em Composite of rotations centered at the same point}: The
  rotations centered at a particular point form a subgroup of the
  group of all self-isometries of $\R^2$. In other words, the
  composite of two rotations centered at the same point is a rotation,
  the identity map is a rotation with any point as center (namely,
  with zero angle of rotation), and the inverse of a rotation is a
  rotation with the same center of rotation. However, the set of {\em
    all} rotations is not a subgroup, as is clear from the next point.
\item {\em Composite of rotations centered at different points}: If
  two rotations centered at different points are composed, the
  composite is typically a rotation about yet a third point, with the
  angle of rotation the sum of the angles. The exception is when the
  angles add up to a multiple of $2\pi$. In that case, the composite
  is a translation. It is easy to convince yourself by using human
  stick figures that the angles of rotation add up.
\item {\em Composite of rotation and translation}: The composite is
  again a rotation with the same angle of rotation, but about a
  different center of rotation.
\item {\em Composite of two reflections}: If the lines of reflection
  are parallel, the composite is a translation by a vector
  perpendicular to both. If the lines of reflection intersect, then
  the composite is a rotation by twice the angle of intersection
  between the lines.
\item {\em Composite of reflection and translation}: This gives rise
  to what is called a {\bf glide reflection}, which is a new type of
  self-isometry of $\R^2$.
\item {\em Composite of reflection and rotation}: This is trickier. It
  could be a reflection or a glide reflection, depending on whether
  the center of rotation lies on the line of reflection.
\end{enumerate}

The upshot is that:

\begin{itemize}
\item The orientation-{\em preserving} self-isometries of $\R^2$ are
  precisely the translations and rotations. Note that these form a
  group.
\item The orientation-{\em reversing} self-isometries of $\R^2$ are
  precisely the reflections and glide reflections. Note that these do
  not form a group, but {\em together} with translations and
  rotations, they form the group of all self-isometries.
\end{itemize}

\subsection{Self-isometries that are linear}

Let's consider self-isometries that are linear, i.e., they fix the
origin. These are subgroups of the group of all
self-isometries. Explicitly:

\begin{itemize}
\item The orientation-{\em preserving} linear self-isometries of
  $\R^2$ are precisely the rotations about the origin, specified by
  the angle of rotation (determined up to additive multiples of
  $2\pi$). Composing two such rotations involves adding the
  corresponding angles. These form a group. This group is denoted
  $SO(2,\R)$ (you don't need to know this!) and is called the {\em
    special orthogonal group} of degree two over the reals.
\item The orientation-{\em reversing} linear self-isometries of $\R^2$
  are precisely the reflections about lines through the origin. These
  do not form a group, but {\em together} with rotations about the
  origin, they form a group. The whole group is denoted $O(2,\R)$ (you
  don't need to know this!) and is called the {\em orthogonal group}
  of degree two over the reals.
\end{itemize}

In general, an affine linear automorphism is a self-isometry if and
only if its linear automorphism part is a self-isometry. In other words:

$$\vec{x} \mapsto A\vec{x} + \vec{b}$$

is a self-isometry if and only if $\vec{x} \mapsto A\vec{x}$ is a
self-isometry.

\section{The case of three dimensions}

The case of three dimensions is somewhat trickier than two dimensions,
but we can still come somewhat close to a classification.

\subsection{Rotations about axes}

The simplest type of orientation-preserving self-isometry is a
rotation about an axis of rotation. Euler proved a theorem (called
{\em Euler's rotation theorem}) that every orientation-preserving
self-isometry that fixes a point must be a rotation about an axis
through that point. In particular, all the orientation-preserving
self-isometries of $\R^3$ that are {\em linear} (in the sense of
fixing the origin) are rotations about axes through the origin.

\subsection{Rotations composed with translations}

We know that translations are orientation-preserving self-isometries
of $\R^n$ for any $n$. So are rotations. We also know that the
self-isometries form a group. Thus, composing a rotation about an axis
with a translation should yield a self-isometry. For $n = 2$, any such
self-isometry would already be a rotation. For $n = 3$, this is no
longer the case. It is possible to have orientation-preserving
self-isometries that are expressible as composites of translations and
rotations but are not translations or rotations themselves. For
instance, a rotation about the $z$-axis followed by a translation
parallel to the $z$-axis works.

\subsection{Reflections about planes and their composites}

A reflection about a plane in $\R^3$ is an orientation-reversing
self-isometry of $\R^3$. For instance, the isometry:

$$(x,y,z) \mapsto (-x,y,z)$$

is a reflection about the $yz$-plane. It is orientation-reversing, and
its matrix is:

$$\left[\begin{matrix} -1 & 0 & 0 \\ 0 & 1 & 0 \\ 0 & 0 & 1 \\\end{matrix}\right]$$

We have not yet seen how to compute the determinant of a general $3
\times 3$ matrix. However, the determinant of a diagonal matrix is
simply the product of the diagonal entries. In this case, the
determinant is $-1$, as it should be, since the reflection is
orientation-reversing but, on account of being a self-isometry, is
{\em volume}-preserving.

A composite of two reflections about different planes is an
orientation-preserving self-isometry. If the planes are not parallel,
this is a rotation about the axis of intersection by twice the angle
of intersection between the planes. For instance, the transformation:

$$(x,y,z) \mapsto (-x,-y,z)$$

corresponds to the diagonal matrix:

$$\left[\begin{matrix} -1 & 0 & 0 \\ 0 & -1 & 0 \\ 0 & 0 & 1 \\\end{matrix}\right]$$

This has determinant $1$. It is both orientation-preserving and
area-preserving. We can also think of it as a rotation by an angle of
$\pi$ about the $z$-axis, i.e., a half-turn about the $z$-axis. The
angle is $\pi$ because the individual planes of reflection are
mutually perpendicular (angle $\pi/2$).

Finally, consider a composite of {\em three} reflections. Consider the
simple case where the reflections are about three mutually
perpendicular planes. An example is:

$$(x,y,z) \mapsto (-x,-y,-z)$$

This corresponds to the diagonal matrix:

$$\left[\begin{matrix} -1 & 0 & 0 \\ 0 & -1 & 0 \\ 0 & 0 & -1 \\\end{matrix}\right]$$

The linear transformation here is orientation-reversing on account of
being a composite of an odd number of reflections. It is a
self-isometry, so the determinant should be $-1$, and indeed, the
determinant is $-1$.

More generally, we could have a map of the form:

$$\left[\begin{matrix} -1 & 0 & 0 \\ 0 & \cos \theta & -\sin \theta \\ 0 & \sin \theta & \cos \theta \\\end{matrix}\right]$$

This map reflects the $x$-coordinate and performs a rotation by
$\theta$ on the $yz$-plane.

\section{Where we're hoping to go with this}

In the future, we will build on what we have learned so far in the
following ways:

\begin{itemize}
\item We will understand the procedure for composing linear
  transformations (or more generally affine linear transformations)
  purely algebraically, i.e., in terms of their description using
  matrices and vectors.
\item We will understand criteria for looking at a linear
  transformation algebraically to determine whether it is a
  self-isometry, self-homothety, orientation-preserving, and/or
  area-preserving.
\item We will understand more about the group structure of various
  kinds of groups of linear transformations and affine linear
  transformations.
\end{itemize}


\end{document}

