\documentclass[10pt]{amsart}
\usepackage{fullpage,hyperref,vipul, graphicx}
\title{Review sheet for midterm 2}
\author{Math 153, Section 55 (Vipul Naik)}

\begin{document}
\maketitle

{\bf To maximize efficiency, please bring a copy (print or readable
electronic) of this review sheet AND the previous review sheet to the
review session.}

This review sheet does {\em not} repeat review of material in the
midterm 1 syllabus, although some of the ``Quickly'' stuff does not
overlap. So, please go through the midterm 1 review sheet too. In the
review session, we will concentrate on the midterm 2 review sheet, but
we can cover selected material fromthe midterm 1 review sheet.

The document is arranged as follows. The initial sections/subsections
correspond to topics. Each subsection has two sets of points,
``Words'' which includes basic theory and definitions, and ``Actions''
which provides information on strategies for specific problem
types.

There may be parts of the lectures that are not mentioned in this
review sheet. Sometimes, these are the parts of lectures that were
optional and were included for ``fun'' purposes, such as an
explanation of present value, the separation of secular trends from
seasonal trends, etc. The review sheet is an attempt to strike a
balance between brevity and relevance on the one hand and
comprehensiveness on the other. If you're keen on more, go back to the
lecture notes. If you're keen on less, be selective in what you read.

To maximize efficiency in the review session, here is what I
suggest. Go through all the lists of points. For each point, make sure
you understand it by jotting down a relevant example or illustration
or providing a brief justification. If you have difficulty, go back to
the lecture notes and read them in detail. You might also want to look
at more worked examples in the book, and check out homework and quiz
problems.

If everybody is on top of the basic material, we will go very quickly
over Sections 1--5 and then spend the rest of the time doing practice
of integration problems from the integration worksheet and mastering
all the skills in the ``Quickly'' list.

\section{Left-overs from integration}

\subsection{Integrating radicals}
Words ...

\begin{enumerate}
\item Expressions of the form $a^2 + x^2$ (with $a > 0$) in the
  denominator or under the radical sign suggest the substitution
  $\theta = \arctan(x/a)$. With this substitution, $x = a \tan
  \theta$, $dx = a \sec^2 \theta \, d\theta$, $a^2 + x^2 = a^2 \sec^2
  \theta$, and $\sqrt{a^2 + x^2} = a \sec \theta$. In the end, when
  substituting back, we use $\theta = \arctan(x/a)$, $\tan \theta =
  x/a$, $\sec \theta = \sqrt{a^2 + x^2}/a$, $\cos \theta = a/\sqrt{a^2
  + x^2}$, and $\sin \theta = x/\sqrt{a^2 + x^2}$. The first sentence
  of substitutions is useful when converting the given integrand into
  a trigonometric integrand. The second sentence is useful when
  converting the integrated answer back at the end. (This latter step
  is unnecessary when we are dealing with a definite integral and we
  transform limits simultaneously).
\item For $a^2 - x^2$ under a squareroot, we have a similar
  substitution $\theta = \arcsin(x/a)$. For $x^2 - a^2$, we take
  $\theta = \arccos(a/x)$. It is useful to work out the forward and
  backward substitutions for these. (See the notes for the details of
  these substitutions). {\em It is strongly suggested that you
  internalize both the forward and the backward substitutions to the
  point where they become automatic. Memorization helps, but you
  should also be able to re-derive things on the spot as the need
  arises.}
\item There is a little subtlety in these substitutions. When we take
  $\theta$ as $\arcsin$, we know that $\cos \theta$ is
  nonnegative. Hence, when we simplify $\sqrt{a^2 - x^2}$, we get
  $\sqrt{a^2\cos^2 \theta}$. Because by assumption $a$ is positive,
  and because $\cos \theta$ is nonnegative, we can write the answer as
  $a\cos \theta$. In other words, we know how exactly we can lift off
  the squareroot. Something similar happens when we are dealing with
  the tangent and secant functions: secant is nonnegative on the range
  of arc tangent. Unfortunately, tangent is {\em not} nonnegative on
  the entire range of arc secant, so we need to actually look at the
  region where we are carrying out the integration. In case both the
  upper and lower bounds of integration are greater than $a$, we know
  that we will in fact get $\tan \theta$.
\end{enumerate}

{\em Note}: Some of you may find it useful to draw right triangles, as
suggested in the book, if reading trigonometric ratios off triangles
is easier for you than algebraic manipulation of trigonometric
expressions.

Actions ...

\begin{enumerate}
\item Trigonometric substitutions allow us to integrate things like
  $x^m(a^2 + x^2)^{n/2}$. However, some special cases of these can be
  integrated without resort to trigonometric substitutions. For
  instance, when $n$ is a nonnegative even integer, this is a sum of
  powers of $x$ and can be integrated term wise. Also, if $m$ is odd,
  we can do a $u$-substitution with $u = a^2 + x^2$.
\item Similar remarks apply to expressions involving $\sqrt{a^2 -
  x^2}$ and $\sqrt{x^2 - a^2}$.
\item To apply this or similar techniques to more general quadratics,
  we need to use a technique known as {\em completing the
  square}. Here, we rewrite:

  $$Ax^2 + Bx + C = A(x + (B/2A))^2 + (C - B^2/4A)$$

  The special case where $A = 1$ is given by:

  $$x^2 + Bx + C = (x + (B/2))^2 + (C - (B/2)^2)$$

  Note that the left-over constant term after completing the square is
  $-D/4A$ where $D$ is the discriminant of the quadratic
  polynomial. In the case $A = 1$, when the polynomial has positive
  discriminant, this left-over term is negative, whereas when the
  polynomial has negative discriminant, this left-over term is
  positive. In the latter case, we can write it as the square of
  something. We would thus have written our original polynomial as $(x
  - \beta)^2 + \gamma^2$, whereupon we can make the substitution
  $\theta = \arctan((x - \beta)/\gamma)$ (or directly apply the
  integration formula).
\end{enumerate}

\subsection{Partial fractions}

Words ...

\begin{enumerate}
\item For most practical purposes, we can study {\em monic}
  polynomials instead of arbitrary polynomials. A monic polynomial is
  a polynomial whose leading coefficient is $1$. The reason we can
  restrict attention to monic polynomials is that any nonzero
  polynomial can be expressed as a nonzero constant times a monic
  polynomial.
\item A nonconstant monic quadratic polynomial is irreducible (i.e.,
  cannot be expressed as a product of polynomials of smaller degree)
  if and only if it has negative discriminant.
\item Every nonconstant monic polynomial with real coefficients is a
  product of monic linear polynomials and irreducible monic
  quadratics, and this factorization is unique. Thus, all irreducible
  monic polynomials are either linear or quadratics with negative
  discriminant.
\item The partial fractions approach breaks up any rational function
  as the sum of a polynomial and rational functions of the form
  $R/Q^k$ where $Q$ is a monic irreducible factor of the original
  denominator and $R$ is a polynomial of degree strictly less than the
  degree of $Q$.
\item Each of these partial fraction pieces is easy to integrate. The
  case where $Q$ is linear, it is of the form $x - \alpha$, and the
  numerator is a constant, so this is a straightforward power
  integration. In the case where $Q$ is quadratic, we break $R$ as the
  sum of a constant and the derivative of $Q$. The constant part is
  handled by a trigonometric substitution, and the derivative of $Q$
  part is handled by the $u$-substitution $u = Q$.
\item The partial fractions approach shows that every rational
  function can be integrated, and we obtain an antiderivative that
  involves $\ln$ (evaluated at some linear function of $x$), $\arctan$
  (again, evaluated at some linear function of $x$), and other rational
  functions.
\item Using the partial fractions approach and the equivalence of
  repeated integrability with the integrability of $x$ times a
  function, we can show that any rational function can be {\em
  repeatedly} integrated, with the final answer in terms of $\arctan$,
  $\ln$, and rational functions.
\end{enumerate}

Actions ... Please go through the notes on partial fractions as well
as the discussion of these in the book. We here list only some salient
points:

\begin{enumerate}
\item Before beginning, make the denominator monic, and use the
  Euclidean algorithm to reduce to a problem where the degree of the
  numerator is less than the degree of the denominator.
\item The general approach is to first factorize the denominator and
  then break it up into partial fractions with unknown numerators. The
  coefficients of the numerator need to be determined. One way of
  doing this is to take a common denominator, multiply out, compare
  coefficients, and solve the resultant system of linear equations.
\item Instead of equating coefficients, we can also use a strategy of
  plugging in values. We plug in values so that a large number of the
  expressions that we are evaluating become zero.
\item In particular, if we want to write:

  $$\frac{r(x)}{(x - \alpha_1)(x - \alpha_2) \dots (x - \alpha_n)} = \frac{c_1}{x - \alpha_1} + \dots + \frac{c_n}{x - \alpha_n}$$

  where all the $\alpha_i$s are distinct and the degree of $r$ is less
  than $n$, then we get:

  $$c_i = \frac{r(\alpha_i)}{\text{product of $\alpha_i - \alpha_j$, all $j \ne i$}}$$

  We can use this to very rapidly write any fraction with denominator
  a product of distinct linear factors in terms of partial fractions,
  and then integrate it.
\item To handle:

  $$\frac{r(x) \, dx}{(q(x))^k}$$

  where $q$ is an irreducible quadratic, we do repeated division,
  taking quotients and remainders, and obtain the result in terms of
  partial fractions.
\item A thorough understanding of the partial fractions approach
  should allow you to predict, simply by looking at a rational
  function, whether the antiderivative expression for it will be (i) a
  rational function, (ii) something involving rational functions and
  $\arctan$, (iii) something involving rational functions and $\ln$,
  or (iv) something involving rational functions, $\arctan$, and
  $\ln$. For some practice of these, refer to the integration quiz.
\end{enumerate}

\subsection{Improper integrals}

Words ...

\begin{enumerate}
\item The integral $\int_a^\infty f(x) \, dx$ is defined as the limit
  $\lim_{L \to \infty} \int_a^L f(x) \, dx$. If $F$ is an
  antiderivative of $f$, this equals $\lim_{L \to \infty} F(L) - F(a)$.
\item The integral $\int_{-\infty}^a f(x) \, dx$ is defined as the
  limit $\lim_{L \to -\infty} \int_L^a f(x) \, dx$. If $F$ is an
  antiderivative of $f$, this equals $F(a) - \lim_{L \to -\infty} F(L)$.
\item The integral $\int_{-\infty}^\infty f(x) \, dx$ is defined as a
  double limit, where the upper limit of integration is limited to
  infinity, while the lower limit of integration is limited to
  negative infinity. If $F$ is an antiderivative of $f$, this equals
  $\lim_{L \to \infty} F(L) - \lim_{M \to -\infty} F(M)$.
\item Another kind of improper integral occurs where the function is
  integrated over an interval and is not defined at an endpoint of the
  interval. Here, we take the limit over intervals of integration
  where the interval gradually tends towards the trouble points. For
  instance, if a function $f$ is to be integrated over $[a,b]$ but $b$
  is a trouble point, we take $\lim_{c \to b} \int_a^c f(x) \, dx$. If
  $F$ is an antiderivative of $f$, then if $F$ extends continuously to
  $b$, this is just equal to $F(b) - F(a)$.
\item In general, if there are multiple trouble points, we first
  partition the interval of integration so that all the trouble points
  are at the partition boundaries. We then use the limiting procedure
  on each piece and add up across the pieces.
\end{enumerate}

Actions ...

\begin{enumerate}
\item The most straightforward way of computing an indefinite integral
  is to compute the corresponding antiderivative and take the
  difference between the upper and lower limits.
\item In some cases, this is either infeasible or terribly messy. In
  these cases, we may use the various other methods for computing
  definite integrals that bypass computing the antiderivative. These
  include the use of symmetry and a combination of $u$-substitution
  plus noticing that after the substitution, the upper and lower
  limits of integration become the same.
\item In yet other cases, taking the limit of the antiderivative may
  be hard, and we may need to use all the techniques discussed in
  preceding subsections for computing this antiderivative.
\end{enumerate}

\section{Differential equations}

\subsection{Solving differential equations at large}

Words ...

\begin{enumerate}
\item A differential equation with dependent variable $y$ and
  independent variable $x$ is something of the form
  $F(x,y,y',y'',\dots) = 0$.
\item The {\em order} of a differential equation is the {\em largest}
  $k$ for which the $k^{th}$ derivative appears in the differential
  equation. In particular, a {\em first-order differential equation}
  only involves $x$, $y$, and $y'$, and does not involve $y''$ or
  higher derivatives. A {\em second-order differential equation} only
  involves $x$, $y$, $y'$, and $y''$.
\item A {\em polynomial differential equation} is one where $F$ looks
  like a polynomial in $y$ and its derivatives. A {\em linear
  differential equation} is a differential equation of the form:

  $$f_k(x)y^{(k)} + \dots + f_1(x)y' + f_0(x)y = g(x)$$

  We can clear out the coefficient of $y^{(k)}$ by dividing throughout
  by $f_k(x)$. The {\em homogeneous} case is where the right side is
  zero.
\item A {\em particular solution} is a relation $R(x,y) = 0$ that,
  when plugged into the differential equation, satisfies it. (Here,
  higher derivatives are computed using {\em implicit
  differentiation}). A particular solution in {\em functional form} is
  one where we explicitly find a function $f$ with $y = f(x)$ that
  satisfies the differential equation.
\item A {\em solution family} is a family with one or more parameters
  such that for every permissible value of the parameter, we obtain a
  particular solution.
\item The {\em general solution} is a solution family that contains
  all particular solutions.
\item A general principle is that the number of freely varying
  parameters in the general solution, also described as the number of
  {\em degrees of freedom}, equals the order of the original
  differential equation. The reason for this is roughly that the
  number of integrations we do that introduce new degrees of freedom
  equals the order of the differential equation.
\item An {\em autonomous} differential equation is a differential
  equation where the independent variable does not appear explicitly
  (except as the thing in terms of which differentiation is carried
  out). The independent variable can be thought of as {\em
  time}. Autonomous differential equations have the property that any
  time translate of a solution is also a solution. This property is
  found in most physical laws, and essentially states that the
  formulation of the physical law does not depend on when we started
  measuring time, i.e., there is no natural time origin.
\item To solve a second-order differential equation we usually do a
  substitution to break it up into solving two first-order
  differential equations.
\item Of first-order differential equations, there are two broad
  classes that we know how to solve: {\em separable} differential
  equations and {\em linear} differential equations. For the latter
  case, the solution method isolates $y$ as a function of $x$. In the
  former case, we can get a mixed bag situation.
\end{enumerate}

Actions ...

\begin{enumerate}

\item The separable case is where we have $y' = f(x)g(y)$. In this
  case, we rearrange to obtain $\int dy/g(y) = \int f(x) \, dx$, and
  integrate both sides. We need to put the $+C$ on only one side,
  because additive constants emanating from both integrals can be
  combined into one additive constant.
\item In the autonomous separable case, we have $dy/dt = g(y)$, and we
  integrate to obtain $\int dy/g(y) = \int dt$. This is the case that
  arises when we look at the logistic equation and its many variants.
\item In the linear case, we have $y' + p(x)y = q(x)$ (after dividing
  out by any coefficient of $y'$). Let $H(x) = \int p(x) \, dx$. The
  integrating factor that we choose is $e^{H(x)}$. When we multiply by
  this integrating factor the left side becomes the derivative of
  $ye^{H(x)}$. Thus, we obtain:

  $$y = e^{-H(x)} \int q(x)e^{H(x)} \, dx$$

  Note that the $+C$ arises in the {\em inner} integral, so the
  general solution is a particular solution plus $Ce^{-H(x)}$.
\item When solving differential equations (particularly the separable
  case) we often get a solution involving logarithms. In some cases,
  it may be useful to {\em exponentiate} both sides. When we do so,
  the original additive constant $C$ arising from indefinite
  integration becomes a multiplicative constant $e^C$. We can also
  absorb {\em sign uncertainty} into it and define a new constant $k =
  e^C\operatorname{sgn}(y)$ to get the answer in terms of a sign
  expression.
\item In a similar vein to the above, if our answer involves an
  inverse trigonometric function, we can apply the trigonometric
  function to both sides. In this case, the additive constant {\em
  sticks inside}. For instance, if we get $\arctan y = x + C$, then
  applying $\tan$ to both sides yields $y = \tan(x + C)$. To simplify
  this further (if we so desire), we need to use the angle sum
  formula. The other major caveat that we need to bear in mind is that
  there is a {\em loss of information} when we apply the trigonometric
  substituion to both sides, because an inverse trigonometric function
  value is constrained to a particular range. This constraint needs to
  be kept track of separately.
\item In some cases, before exponentiating or applying the inverse
  trigonometric function, it might help to use the initial value
  condition to pin down the freely varying parameters (see the next
  subsection).
\end{enumerate}

\subsection{Graphical interpretation and initial value problems}

Words ...

\begin{enumerate}
\item Any {\em particular solution} (whether expressed with $y$ as an
  explicit function of $x$ or in terms of a relation between $x$ and
  $y$) can be plotted as a curve in the $xy$-plane. When it is an
  explicit function, this is the {\em graph of a function} --
  otherwise, it's just the set of points satisfying the relation. This
  picture in the plane is called an {\em integral curve} or a {\em
  solution curve} for the differential equation.
\item The {\em general solution} is thus a picture which has all the
  particular solutions marked.
\item Since solving a $k^{th}$ order differential equation introduces
  $k$ degrees of freedom, we expect that to pin down a unique
  solution, we need $k$ pieces of information. In particular, to
  choose one particular solution for a first-order differential
  equation, we need (by and large) one piece of information. In an
  {\em initial value problem}, this is provided by specifying an
  initial value, which is one point $(x_0,y_0)$ on the particular
  solution curve.
\item Geometrically, we expect that the solution family to a
  first-order differential equation has one real parameter and that,
  except in some degenerate cases, knowing one point on the curve
  determines the curve. In other words we expect that by and large,
  the solution curves do not intersect.
\item For higher-order differential equations, on the other hand, we
  expect that even after knowing one point on the curve, we have
  pinned down only one of many degrees of freedom, and we still have
  solution families to deal with rather than isolated solutions. More
  information, such as information about higher derivatives, or
  information about the curve passing through other points, is
  desirable.
\end{enumerate}

Actions ... Nothing really, except that we plug in the initial value
condition to pin down the constants.

\section{The least upper bound axiom}

Words ...

\begin{enumerate}
\item The real numbers satisfy the {\em least upper bound property}:
  any nonempty subset of the set of real numbers that is bounded from
  above has a least upper bound {\em that is also a real number}. This
  property does {\em not} hold if we replace the real numbers by the
  rational numbers (i.e., the least upper bound of a set of rationals
  bounded from above exists as a real number, but need not be a
  rational number).
\item The real numbers satisfy the {\em greatest lower bound
  property}: any nonempty subset of the set of real numbers that is
  bounded from below has a greatest lower bound {\em that is also a
  real number}. This property again does {\em not} hold if we replace
  the real numbers by the rational numbers.
\item We can prove the greatest lower bound property using the least
  upper bound property. There are two proofs of this. One of these
  proofs involve {\em reflection}: replacing a set by its set of
  negatives. The other proof, which is there in the book, is also
  worth going through. Please go through it. I'll go through it in
  review session. You will not be asked the proof in the test, but it
  may be helpful for multiple choice questions and other conceptually
  based problems.
\item The natural numbers satisfy a property that is somewhat similar
  to the greatest lower bound property for the reals, but stronger:
  any nonempty subset of the set of natural numbers has a least
  element. This is equivalent to the {\em principle of mathematical
  induction}.
\item If a nonempty subset of the real numbers has a {\em maximum}
  element, then that element is also the least upper bound of the
  set. Conversely, if the least upper bound of a set is in the set,
  then that is also the maximum element of the set.
\item If a nonempty subset of the real numbers has a {\em minimum}
  element, then that element is also the greatest lower bound of the
  set. Conversely, if the greatest lower bound of a set is in the set,
  then that is also the minimum element of the set.
\item A nonempty finite subset always has a maximum and a minimum
  element. Thus, its greatest lower bound and least upper bound are
  both in the set.
\item For an interval with lower endpoint $a$ and upper endpoint $b$,
  the least upper bound is $b$ and the greatest lower bound is
  $a$. Note that this holds for all the four possibilities for the
  interval: $[a,b]$, $(a,b)$, $[a,b)$, and $(a,b]$.
\item If $T$ is a nonempty subset of a nonempty bounded subset $S$ of
  $\R$, any lower bound for $S$ remains a lower bound for $T$ and any
  upper bound for $S$ remains an upper bound for $T$. However, we may
  have an upper bound for $T$ that is {\em not} an upper bound for
  $S$. Similarly, we may have a lower bound for $T$ that is {\em not}
  a lower bound for $S$. Thus, the least upper bound for $T$ is $\le$
  the least upper bound for $S$, and the greatest lower bound for $T$
  is $\ge$ the greatest lower bound for $S$.
\item A set does {\em not} have an upper bound if and only if it has
  arbitrarily large elements. Similarly, a set does {\em not} have a
  lower bound if and only if it has arbitrarily small elements (i.e.,
  negative elements of arbitrarily large magnitude).
\item If $M$ is the least upper bound of a nonempty subset $S$ of
  $\R$, then, for every $\epsilon > 0$, $S$ has a nonempty
  intersection with the interval $(M - \epsilon,M]$. In particular, if
  $M \notin S$, then $S$ has a nonempty intersection with the interval
  $(M - \epsilon, M)$. (See also the analogous theorem for greatest
  lower bounds, which is Theorem 11.1.4 in the book).
\end{enumerate}

Actions ...

\begin{enumerate}
\item To compute the greatest lower bound and least upper bound of a
  set, we first need to compute the set. Finding the set as a union of
  intervals is often useful.
\item Given a set $S$, we can construct corresponding sets such as $S
  + \lambda$ (translation), $-S$ (reflection about $0$), $f(S)$ (image
  of $S$ under a function $f$), and $\operatorname{abs}(S)$ (the set
  of absolute values of elements of $S$, i.e., folding about
  $0$). Please review the results that relate bounds for $S$ with
  bounds on these corresponding sets.
\end{enumerate}

\section{Sequences of reals}

\subsection{Sequencs: basics}
Words ...

\begin{enumerate}
\item A sequence in a set is a function from $\N$ to that set.
\item A sequence of reals can be described in three ways: as an
  ordered list of real numbers (where we write only the first few
  members due to space and time considerations), as a closed form
  expression for the general term of the sequence (i.e., thinking of
  it as a function), and in terms of a recurrence relation.
\item The {\em range} of a sequence is the set of values that it
  takes. The range of a sequence differs from the sequence in the
  following two senses: (i) it ignores repetition (ii) it ignores the
  ordering or sequencing of the elements.
\item There are many properties that we can talk of in the context of
  sequences: increasing, decreasing, non-increasing, non-decreasing,
  monotonic, constant, periodic, bounded from below, bounded from
  above, and bounded. Note that the boundedness-related properties
  depend {\em only} on the range of the sequence, whereas the other
  properties depend on the sequence as a whole.
\item For any property $p$ that can be evaluated on each sequence, we
  can talk of the property {\em eventually} $p$, which means that some
  left shift of the sequence has the property $p$. For instance, we
  can talk of eventually increasing, eventually decreasing, eventually
  constant, eventually monotonic, and eventually periodic.
\item Eventually bounded is the same as bounded.
\item There are various operations we can do on sequences similar to
  the corresponding operations on functions: add, subtract, multiply,
  divide, multiply by a scalar, and compose with a function defined on
  the range of the sequence.
\item We can do left shifts, right shifts (though this requires us to
  throw in more new terms), splicing, and other fancy operations.
\item If a sequence is defined recursively, i.e., using a recurrence
  relation, then we need to separately specify initial values. The
  number of initial values that we need to specify depends on how far
  back the recurrence relation reaches. This is related both to the
  principle of mathematical induction and the idea of free parameters
  and initial value specifications for differential equations.
\item We can define a discrete derivative, called the {\em forward
  difference operator}, defined as follows: for a sequence $f:\N \to
  \R$, the forward difference operator is $(\Delta f)(n) = f(n + 1) -
  f(n)$. This is analogous to the derivative of a continuous function.
\item The {\em integration equivalent} for the forward difference
  operator is the summation operator. (Try to find the formula for
  this in the notes/class discussion).
\item The forward difference operator behaves analogously to
  differentiation (though the formulas differ) for constants,
  polynomial sequences, and periodic sequences. (Please see the notes
  for a list of points).
\item Periodic sequences can be defined using a case-wise definition
  based on the remainder modulo the period. They can alternatively be
  defined as combinations of trigonometric functions. In the special
  case of a period $2$, we can also express in terms of $(-1)^n$.
\item A sequence with periodic derivative can be expressed as the sum
  (pointwise) of a linear sequence and a periodic sequence.
\end{enumerate}

\subsection{Continuous-discrete interplay}

Words ...

\begin{enumerate}
\item Given a function on $\R$, we can restrict the function to $\N$
  and obtain a sequence. This restriction is unique.
\item Conversely, given a sequence, i.e., a function on $\N$, we can
  extend it to a continuous function on $\R$. However, the extension
  is not unique, and there are a lot of different ways of
  extending. If the sequence is described by means of a nice closed
  form functional expression, we may be able to extend it by
  considering that functional expression for all real numbers.
\item Usually, information about the function on the reals gives us
  corresponding information about the corresponding sequence, but we
  cannot get information in the reverse direction that easily. For
  instance, an increasing function gives an increasing sequence, but
  increasing sequences can arise from functions that are not
  increasing. A decreasing function gives a decreasing sequence, a
  monotonic function gives a monotonic sequence, and a bounded
  function gives a bounded sequence.
\item A function with integer period gives a periodic sequence.
\item The mean value theorem relates the derivative of a function to
  the discrete derivative (i.e., forward difference operator) of the
  corresponding sequence.
\item We can define a notion of concave up and concave down for
  sequences based on the second discrete derivative. If a function is
  concave up, so is the corresponding sequence. If the function is
  concave down, so is the corresponding sequence.
\end{enumerate}

\section{Limit computation techniques}

Words ...

\begin{enumerate}
\item L'H\^{o}pital's rule for $0/0$ form: Consider $\lim_{x \to c}
  f(x)/g(x)$ where $\lim_{x \to c} f(x) = \lim_{x \to c} g(x) = 0$. If
  both $f$ and $g$ are differentiable around $c$, then this limit is
  equal to $\lim_{x \to c} f'(x)/g'(x)$.
\item Analogous statements hold for one-sided limits and in the case
  $c = \pm \infty$.
\item The $0/0$ form LH rule {\em cannot} be applied if the numerator
  does not limit to zero or if the denominator does not limit to zero.
\item L'H\^{o}pital's rule for $\infty/\infty$ form: Consider $\lim_{x
  \to c} f(x)/g(x)$ where $\lim_{x \to c} f(x) = \pm \infty$ and
  $\lim_{x \to c} g(x) = \pm \infty$. If both $f$ and $g$ are
  differentiable around $c$, then this limit is equal to $\lim_{x \to
  c} f'(x)/g'(x)$.
\item Analogous statements hold for one-sided limits and for $c = \pm
  \infty$.
\item For a function $f$ having a zero at $c$, the {\em order} of the
  zero at $c$ is the least upper bound of $\beta$ such that $\lim_{x
  \to c} |f(x)|/|(x - c)|^{\beta} = 0$. At this least upper bound, the
  limit is usually finite and nonzero. For larger values of $\beta$,
  the limit is undefined.
\item Given a quotient $f/g$ for which we need to calculate the limit
  at $c$, the limit is zero if the order of the zero for $f$ is
  greater than the order for $g$, and undefined if the order of the
  zero for $f$ is less than the order for $g$. When the orders are the
  same, the limit could potentially be a finite nonzero number.
\end{enumerate}

Actions ...

\begin{enumerate}
\item For polynomial functions and other continuous functions, we can
  calculate the limit at a point by evaluating at the point. For
  rational functions, we can cancel common factors between the
  numerator and the denominator till one of them becomes nonzero at
  the point.
\item There is a bunch of basic limits that translate to saying that
  for the following functions $f$: $f(0) = 0$ and $f'(0) = 1$, which
  is equivalent to saying that $\lim_{x \to 0} f(x)/x = 1$. These
  functions are $\sin x$, $x \mapsto \ln(1 + x)$, $x \mapsto e^x - 1$,
  $x \mapsto \tan x$, $x \mapsto \arcsin x$, and $x \mapsto \arctan
  x$.
\item For all the functions $f$ of the above kind (that we call {\em
  strippable}), the following is true: in any multiplicative
  situation, if the input to the function goes to zero in the limit,
  the function can be stripped off.
\item Two other basic limits are $\lim_{x \to 0} (1 - \cos x)/x^2 =
  1/2$ and $\lim_{x \to 0} (x - \sin x)/x^3 = 1/6$. These can be
  obtained using the LH rule.
\item Typically, to compute limits, we can combine the LH rule,
  stripping, and removing multiplicative components that we can
  calculate directly.
\item Applying the LH rule $0/0$ form pushes the orders of both the
  numerator and the denominator down by one.
\item It is also useful to remember that logarithmic functions are
  dominated by polynomial functions, which in turn are dominated by
  exponential functions. These facts can be seen in various ways,
  including the LH rule.
\end{enumerate}

\section{Tricky topics}

\subsection{Important classes of functions outside of the elementary world}

{\em This is not part of the executive summary of any of the lecture
notes, but is related to some of the homework problems and the
challenge problem.}

We note here three such classes of functions:

\begin{enumerate}
\item The function arising as an integral of $\exp(-x^2)$. All the
  following functions have antiderivatives expressible in terms of
  this antiderivative and elementary functions: 

  \begin{itemize}
  \item $x^{2k}\exp(-mx^2)$ where $k$ is a positive integer and $m$ is
    a positive real: The idea here is to use integration by parts,
    taking $x^{2k - 1}$ as the part to differentiate and
    $x\exp(-mx^2)$ as the part to integrate. The problem reduces to
    $x^{2k - 2}\exp(-mx^2)$. Repeat.
  \item $x^{m + (1/2)}\exp(-x)$ where $m$ is an integer (not
    necessarily positive. Putting $u = \sqrt{x}$ allows us to go back
    and forth with $\exp(-x^2)$. Note that the $x^{m + (1/2)}\exp(-x)$
    are related to each other for different values of $m$ via
    integration by parts: take $\exp(-x)$ as the part to integrate.
  \item $\exp(x^{-2/(2m + 1)}$ where $m$ is a positive integer. Put $u
    = x^{1/(2m + 1)}$ to go back and forth.
  \end{itemize}

  In particular, in applicable cases, we can calculate the {\em
  improper definite integral} on all of $\R$ for each of these
  functions based on the fact that:

  $$\int_{-\infty}^\infty \exp(-x^2) = \sqrt{\pi}$$

  In particular, since the function is even, we have:

  $$\int_0^\infty \exp(-x^2) = \frac{\sqrt{\pi}}{2}$$

  We define:

  $$\Gamma(x) := \int_0^\infty t^{x - 1}e^{-t} \, dt$$

  for $x \in (0,\infty)$.

  Based on the fact about the integral of $\exp(-x^2)$, we can
  calculate the value of $\Gamma$ at all half-integers. Note that for
  $x$ a positive integer, $\Gamma(x) = (x - 1)!$ as seen in a homework
  exercise.

\item The function arising as an integral of $1/(\ln x)$, for $x >
  1$. All the following have antiderivatives expressible in terms of
  this antiderivatve and elementary functions:

  \begin{itemize}
  \item $\ln(\ln x)$ for $x > 1$.
  \item $e^x/x$, for $x > 0$. Also, $e^x\ln x$$e^x/x^r$ for all
    positive integers $r$.
  \item $e^{e^x}$ for all $x$.
  \end{itemize}

\item The function arising as an integral of $(\sin x)/x$ (we fill in
  the value of the integrand at $0$ to be $1$). All the following have
  antiderivatives expressible in terms of this: $(\cos x)(\ln x)$, $(1
  - \cos x)/x^2$, $(x - \sin x)/x^3$. The back-and-forth technique is
  integration by parts: one thing gets integrated, the other
  differentiated.

  In particular, given that:

  $$\int_0^\infty \frac{\sin x}{x} \, dx = \frac{\pi}{2}$$

  we can compute the integrals from $0$ to $\infty$ of $(1 - \cos
  x)/x^2$ and $(x - \sin x)/x^3$.
\end{enumerate}

\section{Quickly}

This section lists things you should be able to do quickly.

\subsection{Sequence pattern finding}

You should be able to do the following:

\begin{enumerate}
\item Given the first few terms of an easy sequence, predict the next
  few terms.
\item Describe a sequence using a recurrence relation.
\item Write the general expression (possibly making cases if there's
  periodic-like behavior) for the $n^{th}$ term.
\item Move back and forth between general term descriptions and
  recurrence relation descriptions of sequences.
\end{enumerate}

\subsection{Our common values}

Preferably remember these (or be capable of computing quickly) to at
least one digit. Generally, you will {\em not} be asked to do any
numerical computations using these. In practice, the main way this is
useful is to figure out whether something is positive or negative. For
instance, is $3\sqrt{2} - 4$ positive? What about $e^2 - 8$? Often,
there are other ways of answering such questions, but remembering the
numerical values is a quick and dirty approach.

\begin{enumerate}
\item Square roots of $2$, $3$, $5$, $6$, $7$, $10$.
\item Natural logarithms of $2$, $3$, $5$, $7$, and $10$.
\item Value of $\pi$, $1/\pi$, $\sqrt{\pi}$, and $\pi^2$.
\item Value of $e$, $1/e$.
\item Some relative logarithms, such as $\log_23$ or
  $\log_2(10)$. Although you don't need these values to a significant
  degree of precision, it is useful to have some idea of their
  magnitude.
\end{enumerate}

\subsection{Adding things up: arithmetic}

You should be able to:

\begin{enumerate}
\item Do quick arithmetic involving fractions.
\item Sense when an expression will simplify to $0$.
\item Compute approximate values for square roots of small numbers,
  $\pi$ and its multiples, etc., so that you are able to figure out,
  for instance, whether $\pi/4$ is smaller or bigger than $1$, or two
  integers such that $\sqrt{39}$ is between them.
\item Know or quickly compute small powers of small positive
  integers. This is particularly important for computing definite
  integrals. For instance, to compute $\int_2^3 (x + 1)^3 \, dx$, you
  need to know/compute $3^4$ and $4^4$.
\end{enumerate}

\subsection{Computational algebra}

You should be able to:

\begin{enumerate}
\item Add, subtract, and multiply polynomials.
\item Factorize quadratics or determine that the quadratic cannot be
  factorized.
\item Factorize a cubic if at least one of its factors is a small and
  easy-to-spot number such as $0$, $\pm 1$, $\pm 2$, $\pm 3$. {\em
  This could be an area for potential improvement for many people.}
\item Factorize an even polynomial of degree four. {\em This could be
  an area for potential improvement for many people.}
\item Do polynomial long division.
\item Solve simple inequalities involving polynomial and rational
  functions once you've obtained them in factored form.
\end{enumerate}

\subsection{Computational trigonometry}

You should be able to:

\begin{enumerate}
\item Determine the values of $\sin$, $\cos$, and $\tan$ at multiples
  of $\pi/2$.
\item Determine the intervals where $\sin$ and $\cos$ are positive and
  negative.
\item Remember the formulas for $\sin(\pi \pm x )$ and $\cos(\pi \pm x)$,
  as well as formulas for $\sin(-x)$ and $\cos(-x)$.
\item Recall the values of $\sin$ and $\cos$ at $\pi/6$, $\pi/4$, and
  $\pi/3$, as well as at the corresponding obtuse angles or other
  larger angles.
\item Reverse lookup for these, for instance, you should quickly
  identify the acute angle whose $\sin$ is $1/2$.
\item Formulas for double angles, half angles: $\sin(2x)$, $\cos(2x)$
  in terms of $\sin$ and $\cos$; also the reverse: $\sin^2x$ and
  $\cos^2x$ in terms of $\cos(2x)$.
\item Remember the formulas for $\sin(A + B)$, $\cos(A + B)$, $\sin(A
  - B)$, and $\cos(A - B)$.
\item Convert between products of $\sin$ and $\cos$ functions and
  their sums: for instance, the identity $2\sin A \cos B = \sin(A + B)
  + \sin (A - B)$. You don't have to remember these identities
  separately since they follow from the identities covered in the
  previous point, but you should be comfortable going back and forth.
\end{enumerate}

\subsection{Computational limits}

This includes:

\begin{enumerate}
\item Look at a rational function and determine its limit to $+\infty$
  and $-\infty$ based on heuristics, without thought.
\item Look at something involving a quotient of an exponential and a
  polynomial, or a polynomial and a logarithmic, and predict the
  behavior as we go to $0$ or $\infty$.
\item Know which functions can be stripped from composites.
\item Apply the LH rule mentally multiple times and figure out how
  many times it needs to be applied.
\end{enumerate}

\subsection{Computational differentiation}

You should be able to:

\begin{enumerate}
\item Differentiate a polynomial (written in expanded form) once or
  twice on sight, without rough work.
\item Differentiate sums of powers of $x$ on sight (without rough
  work).
\item Differentiate rational functions with a little thought.
\item Do multiple differentiations of expressions whose derivative
  cycle is periodic, e.g., $a \sin x + b \cos x$ or $a \exp(-x)$.
\item Do multiple differentiations of expressions whose derivative
  cycle is periodic up to constant factors, e.g. $a \exp(mx + b)$ or $a
  \sin(mx + \varphi)$.
\item Differentiate simple composites without rough work (e.g.,
  $\sin(x^3)$).
\item Differentiate $\ln$, $\exp$, and expressions of the form $f^g$
  and $\log_f(g)$.
\end{enumerate}

\subsection{Computational integration}

You should be able to:

\begin{enumerate}
\item Compute the indefinite integral of a polynomial (written in
  expanded form) on sight without rough work.
\item Compute the definite integral of a polynomial with very few
  terms within manageable limits quickly.
\item Compute the indefinite integral of a simple-looking rational
  function and also the definite integral, with a little time,
  including improper integrals.
\item Compute the indefinite integral of a sum of power functions
  quickly.
\item Know that the integral of sine or cosine on any quadrant is $\pm
  1$.
\item Compute the integral of $x \mapsto f(mx)$ if you know how to
  integrate $f$. In particular, integrate things like $(a + bx)^m$.
\item Integrate $\sin$, $\cos$, $\sin^2$, $\cos^2$, $\tan^2$,
  $\sec^2$, $\cot^2$, $\csc^2$, $\sin^3$, $\cos^3$, $\tan^3$,
  $\sec^3$, $\cot^3$, $\csc^3$, and other higher powers of the basic
  trigonometric functions.
\item Integrate on sight things such as $x\sin(x^2)$, getting the
  constants right without much effort.
\item {\em By parts}: Integrate $\int xf(x) \, dx$ on sight if it is
  easy to integrate $f$ twice, without having to think much (the
  answer is $x \int f - \int \int f$). Similarly, do $\int x^2f(x) \,
  dx$ if you know how to integrate $f$ twice.
\item Using the previous point, integrate $\int f(\sqrt{x}) \, dx$
  with minimal stress and effort.
\item {\em By parts}: Integrate $\int f(x) \, dx$ on sight if $xf'(x)$
  is easy to integrate, e.g., $\int \ln x \, dx$.
\item Remember the integrals for formats such as $e^x \cos x$ and $e^x
  \sin x$.
\item If there is an easy choice of $f$ such that $f + f' = g$,
  integrate $\int e^xg(x) \, dx = e^xf(x)$ on sight and similarly
  integrate $\int g(\ln x) \, dx = xf(\ln x)$ on siht.
\end{enumerate}
\subsection{Being observant}

You should be able to look at a function and:

\begin{enumerate}
\item Sense if it is odd (even if nobody pointedly asks you whether it
  is).
\item Sense if it is even (even if nobody asks you whether it is).
\item Sense if it is periodic and find the period (even if nobody asks
  you about the period).
\end{enumerate}

\subsection{Graphing}

You should be able to:

\begin{enumerate}
\item Mentally graph a linear function.
\item Mentally graph a power function $x^r$ (see the list of things to
  remember about power functions). Sample cases for $r$: $1/3$, $2/3$,
  $4/3$, $5/3$, $1/2$, $1$, $2$, $3$, $-1$, $-1/3$ $-2/3$.
\item Graph a piecewise linear function with some thought.
\item Mentally graph a quadratic function (very approximately) --
  figure out conditions under which it crosses the axis etc.
\item Graph a cubic function after ascertaining which of the cases for
  the cubic it falls under.
\item Mentally graph $\sin$ and $\cos$, as well as functions of the $A
  \sin(mx)$ and $A\cos(mx)$.
\item Graph a function of the form linear + trigonometric, after doing
  some quick checking on the derivative.
\item Graph the inverse trigonometric functions $\arctan$, $\arcsin$,
  and $\arccos$.
\end{enumerate}

\subsection{Graphing: transformations}

Given the graph of $f$, you should be able to quickly graph the following:

\begin{enumerate}
\item $f(mx)$, $f(mx + b)$: pre-composition with a linear function;
  how does $m < 0$ differ from $m > 0$?
\item $Af(x) + C$: post-composition with a linear function, how does
  $A > 0$ differ from $A < 0$?
\item $f(|x|)$, $|f(x)|$, $f(x^+)$, and $(f(x))^+$: pre- and
  post-composition with absolute value function and positive part
  fnuction.
\item More slowly: $f(1/x)$, $1/f(x)$, $\ln(|f(x)|)$, $f(\ln|x|)$,
  $\exp(f(x))$, and other popular composites.
\end{enumerate}

\end{document}