\documentclass[10pt]{amsart}

%Packages in use
\usepackage{fullpage, hyperref, vipul, enumerate}

%Title details
\title{Class quiz: September 30: Limits}
\author{Vipul Naik}
%List of new commands

\begin{document}
\maketitle

Your name (print clearly in capital letters): $\underline{\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad}$

Write your answer in the space provided. In the space below, you can
explain your work if you want (this will not affect scoring). I may or
may not get time to look at the work you have done, but it may help
you recall how you arrived at a particular answer.

You are expected to take about one minute per question.

Questions marked with a (*) are questions that are somewhat trickier,
with the probability of getting the question correct being about
$50\%$ or less. For these questions, you are free to discuss the questions
with others while making your attempt.

Questions marked with a (**) are questions where, in a previous
administration of this quiz, a specific incorrect option was chosen by
as many people as or more people than the correct option. For these
questions, you are free to discuss the questions with others while
making your attempt.

\begin{enumerate}

\item (**) We call a function $f$ left continuous on an open interval $I$
  if, for all $a \in I$, $\lim_{x \to a^-} f(x) = f(a)$. Which of the
  following is an example of a function that is left continuous but
  not continuous on $(0,1)$? {\em Last year's performance}: $6/13$ correct

  \begin{enumerate}[(A)]
  \item $f(x) := \lbrace\begin{array}{rl}x, & 0 < x \le 1/2 \\ 2x, & 1/2 < x < 1 \\\end{array}$
  \item $f(x) := \lbrace\begin{array}{rl}x, & 0 < x < 1/2 \\ 2x, & 1/2 \le x < 1 \\\end{array}$
  \item $f(x) := \lbrace\begin{array}{rl}x, & 0 < x \le 1/2 \\ 2x - (1/2), & 1/2 < x < 1 \\\end{array}$
  \item $f(x) := \lbrace\begin{array}{rl}x, & 0 < x < 1/2 \\ 2x - (1/2), & 1/2 \le x < 1 \\\end{array}$
  \item All of the above
  \end{enumerate}

  \vspace{0.1in}
  Your answer: $\underline{\qquad\qquad\qquad\qquad\qquad\qquad\qquad}$
  \vspace{1.5in}
  
\item (**) Suppose $f$ and $g$ are functions $(0,1)$ to $(0,1)$ that
  are both left continuous on $(0,1)$. Which of the following is {\em
    not} guaranteed to be left continuous on $(0,1)$? {\em Last year's
    performance}: $4/13$ correct

  \begin{enumerate}[(A)]
  \item $f + g$, i.e., the function $x \mapsto f(x) + g(x)$
  \item $f - g$, i.e., the function $x \mapsto f(x) - g(x)$
  \item $f \cdot g$, i.e., the function $x \mapsto f(x)g(x)$
  \item $f \circ g$, i.e., the function $x \mapsto f(g(x))$
  \item None of the above, i.e., they are all guaranteed to be left
  continuous functions
  \end{enumerate}

  \vspace{0.1in}
  Your answer: $\underline{\qquad\qquad\qquad\qquad\qquad\qquad\qquad}$
  \vspace{1.5in}

\item (*) Consider the function:

  $$f(x) := \lbrace\begin{array}{rl} x, & x \text{ rational }\\1/x, & x \text{ irrational }\\\end{array}$$

  What is the set of all points at which $f$ is continuous? {\em Last
    year's performance}: $5/13$ correct

  \begin{enumerate}[(A)]
  \item $\{ 0, 1 \}$
  \item $\{ -1,1 \}$
  \item $\{-1,0 \}$
  \item $\{ -1,0,1 \}$
  \item $f$ is continuous everywhere
  \end{enumerate}

  \vspace{0.1in}
  Your answer: $\underline{\qquad\qquad\qquad\qquad\qquad\qquad\qquad}$
  \vspace{1.5in}

\end{enumerate}
\end{document}