\documentclass[10pt]{amsart}

%Packages in use
\usepackage{fullpage, hyperref, vipul, enumerate}

%Title details
\title{Words and pictures: October 26: Calculus in the real world}
\author{Math 152, Section 55 (Vipul Naik)}
%List of new commands

\begin{document}
\maketitle

Your name (print clearly in capital letters): $\underline{\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad}$

This is just a fun quiz.
\begin{enumerate}

\item ``The Reaction Time game is a game used to measure people's
  reaction time. When people start playing the game, their reaction
  time is in the 250-300ms range. We find that repeatedly playing the
  Reaction Time game causes the average reaction time to drop
  (allowing for random fluctuation as well as day-to-day variation on
  account of the person's physiological state). However, it does not
  approach zero but rather converges toward an {\em asymptotic}
  positive value of 150ms.''

  \vspace{1.5in}

\item ``Our website traffic shows significant seasonal variation, with
  seasonal peaks during the US academic year and seasonal troughs
  during the summer vacation. We also have weekly periodic variation,
  with intra-week highs Monday - Thursday and lows Friday -
  Sunday. Controlling for seasonal variation, daily traffic to our
  website is growing by 1000 pageviews per day every year.''

  \vspace{1.5in}

\item ``The instructor was sued under discrimination law because the
  letter grades he assigned on the test were not a {\em non-decreasing
  function} of the students' raw score on the test.''
  
  \vspace{1in}

\item ``Opportunity cost and the reality of trade-offs mean that the
  common welfare will not be maximized if all resources are
  exclusively used in the production of wheat, nor if none
  are. Rather, the optimal amount of resources that ought to be
  devoted to wheat production lies in the middle. It is the middles,
  not the extremes, that are best for the common welfare.''

  \vspace{1in}

\item ``The most pernicious forms of laziness and complacency arise
  when you are at a {\em local maximum} and are therefore unwilling to
  take a temporary dip in order to strive toward the bigger {\em absolute
  maximum}.''

  \vspace{1.5in}

\item ``Every day, the daily sales at my lemonade stand go up by one,
  so the cumulative sales so far are a {\em quadratic function} of the
  number of days I have kept my lemonade stand open.''

  \vspace{1.5in}

\item ``My Facebook friend count continues to grow, but the rate at
  which it is growing is steadily decreasing as I have to spend more
  and more time on work and less and less time on meeting new people
  outside my immediate friend circle.''

  \vspace{1in}
\item ``In small quantities, enzyme $C$ acts as a catalyst for its own
  creation (heard of autocatalysis?). However, once the quantity of
  enzyme $C$ exceeds a certain threshold value, it acts as an
  inhibitor for its own creation. The biochemical reaction that
  produces enzyme $C$ thus begins slowly, speeds up, and then reaches
  an {\em inflection point} after which it starts slowing down and
  eventually tapers off to zero. (Think of the graphs for the quantity
  of $C$ produced and the rate of $C$ produced per unit time).''

  \vspace{1.5in}

\end{enumerate}

\end{document}