\documentclass[10pt]{amsart}

%Packages in use
\usepackage{fullpage, hyperref, vipul, amssymb, setspace, enumerate}

%Title details
\title{Lazard correspondence up to isoclinism: pending items}
\author{Vipul Naik}

%List of new commands
\newcommand{\Skew}{\operatorname{Skew}}
\newcommand{\ad}{\operatorname{ad}}
\begin{document}
\maketitle
\onehalfspacing

{\em Necessary for completeness}:

\begin{enumerate}

\item The definitions of the ``commutator-like map'' and the ``Lie
  bracket-like map'' that have been made in the context of groups need
  to also be made in the context of group extensions. Further, I need
  to write out a verification that the canonical isomorphism between
  the Schur multipliers $M(G)$ and $M(L)$ satisfies a compatibility
  condition with these maps (this is intuitively clear, but the formal
  proof requires some diagram-chasing and I would need some time to
  write it out). This will be necessary for me to demonstrate the
  equivalence of two different interpretations I have provided for the
  Lazard correspondence up to isoclinism. I estimate that this will
  take about 2 hours.

\item I have demonstrated a Lazard correspondence between a group
  $[K,K]$ and a Lie ring $[N,N]$ (where $K$ and $N$ themselves are not
  in Lazard correspondence, but $\sqrt{K}$ and $\sqrt{N}$ are in
  Lazard correspondence). I have to show that under this
  correspondence, particular subgroups of $[K,K]$ correspond to
  particular subrings of $[N,N]$. I have a proof outline, but need to
  write it out clearly. I estimate that this will take about 2 hours.
\end{enumerate}

After making these changes, I need to review the last three sections
carefully to see if they make sense and whether any additional
material needs to be added. I estimate that will take 1-2 hours.

{\em Necessary for full result strength}: It would look bad to have a
thesis that does {\em not} extend the results to their full strength
as below.

\begin{enumerate}

\item I want to relax the assumption that the whole group $G$ needs to
  be $\pi_c$-powered. Instead, I should require only that $G'$ and
  $\operatorname{Inn}(G)$ are $\pi_c$-powered, or even better, that
  only $\operatorname{Inn}(G)$ is $\pi_c$-powered. The proofs are very
  similar, but the notation and language need to be generalized and
  made more complicated. I estimate this will take about 4 hours.
\item I want to formulate my results for the general version of the
  Lazard correspondence ($3$-local). Currently, the statement has been
  formulated in the $3$-local case, but the full proof of existence
  has been written out only for the global case. For the $3$-local
  case, I still have a definition of the correspondence up to
  isoclinism but I do not have a proof that there always exist groups
  in correspondence up to isoclinism with a given Lie ring, or vice
  versa. The problem with the $3$-local case is that the corresponding
  free group and Lie ring and no longer the usual free nilpotent group
  and Lie ring. Thus, I need to introduce new notation, despite the
  proofs being structurally similar. I am unclear on how much
  additional work this will take, or whether it would be worthwhile to
  do this right now. I estimate {\em at least} 5 hours. I will have a
  better idea of how much time is needed after spending about 2 hours
  on this.
\end{enumerate}

\end{document}
