\documentclass[10pt]{amsart}

%Packages in use
\usepackage{fullpage, hyperref, vipul, amssymb, setspace, enumerate}

%Title details
\title{Lazard correspondence up to isoclinism: changes}
\author{Vipul Naik}

%List of new commands
\newcommand{\Skew}{\operatorname{Skew}}
\newcommand{\ad}{\operatorname{ad}}
\begin{document}
\maketitle
\onehalfspacing

\begin{itemize}
\item No changes to Section 1 (introduction).
\item Added Section 2 (outline, already seen).
\item Added Section 3 (abelian Lie correspondence).
\item Sections 4 to 14 same as before (corresponds to earlier Sections
  2 - 12).
\item New sections 15 and 16 split from later section.
\item Sections 17 and 18 same as before. Some proofs here that involve
  tedious induction omit steps by citing analogous steps earlier.
\item Sections 19 corresponds to last section you read from the August
  25 draft.
\item Section 20 is a new section with the Lie ring analogues of
  Section 19.
\item Sections 21 and 22 describe the Baer correspondence. Content is
  similar but presentation has been updated.
\item Section 23 describes the {\em Baer correspondence up to
  isoclinism}. Content is similar but presentation has been updated.
\item Sections 24 and 25 are similar to Sections 17 and 18 of the
  August 25 version, with minor fixes.
\item Sections 26-30 provide an overview of the Malcev and Lazard
  correspondence. They have been updated since the August 25 version,
  but I think you didn't get to these sections in the August 25
  version.
\item Section 31 has important results about prime divisors of the
  denominators in formulas relating the commutator and Lie bracket.
\item Sections 32-34 apply the result of Section 31 to establish the
  Lazard correspondence up to isoclinism.
\end{itemize}

Possible typos (I have tried to fix all these, but I might have missed
a few):

\begin{itemize}
\item I changed notation in Section 31 to use class $c + 1$ instead of
  class $c$ for consistency with later sections. It may be possible
  that I did not update the notation consistently.
\item I switched notation from $\pi$ to $\pi_c$ to describe the set of
  primes less than or equal to $c$, but there may be cases where I
  omitted updating the notation.
\item My compiler says there were no undefined references, so you
  should not see ``??'' anywhere. However, there may still be wrong
  references (reference to the incorrect section or theorem number)
  because I moved theorems around a lot at the last minute.
\end{itemize}

Possibly unclear:

\begin{itemize}
\item Some category theory terminology may not be adequately explained
  at the point of use (I have tried to fix this).
\item At a few places, I have used the word ``canonical'' to describe
  an isomorphism and the meaning is vague.
\item The scope of some notation borrowed from Khukhro (that is being
  carried across sections) may be difficult for readers to keep in mind.
\item Looking at internal references is often tedious, and it may be
  the case that I should more clearly recall the statement of a
  theorem or the contents of a section that I am referencing.
\end{itemize}
\end{document}
