\documentclass{ucetd}

%Packages in use
%\usepackage{fullpage}
%\usepackage{hyperref}
\usepackage{vipul}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace, enumerate}
\usepackage{appendix}

%Title details
\title{Lazard correspondence up to isoclinism}
\author{Vipul Naik}
\department{Mathematics}
\division{The Physical Sciences}
\degree{Doctor of Philosophy}
\date{December 2013}

%% Use these commands to set a dedication and epigraph text
%\dedication{God}
%\epigraph{Something cheesy}


%List of new commands
\newcommand{\Skew}{\operatorname{Skew}}
\newcommand{\ad}{\operatorname{ad}}
\newcommand{\url}[1]{#1}
\newcommand{\urlfake}[1]{}
\begin{document}
%% Basic setup commands
% If you don't want a title page comment out the next line and uncomment the line after it:
\maketitle
%\omittitle

% These lines can be commented out to disable the copyright/dedication/epigraph pages
%\makecopyright
%\makededication
%\makeepigraph


%% Make the various tables of contents
\tableofcontents
%\listoffigures
\listoftables

\abstract

In this thesis we generalize the Lazard correspondence, introduced by
Lazard in \cite{Lazardsoriginal}, to a correspondence {\em up to
  isoclinism}. The original Lazard correspondence is a correspondence
between some groups and some Lie rings. The Lazard correspondence up
to isoclinism is a correspondence between some equivalence classes of
groups and some equivalence classes of Lie rings, where the
equivalence relation on both sides is isoclinism. By relaxing the
objects on both sides of the correspondence to equivalence classes up
to isoclinism, we are able to generalize the domain of the
correspondence somewhat. An overview of our proof strategy is in
Section \ref{sec:outline}, and our final results are described in
Section \ref{sec:lcuti}.

A typical application of the original Lazard correspondence is the
situation where, for some prime $p$, the group is a finite $p$-group
of nilpotency class at most $p - 1$ and the Lie ring is a finite
$p$-Lie ring\footnote{This means that the additive group of the Lie
  ring is a finite $p$-group. In other words, the Lie ring is a Lie
  algebra over $\Z/p^k\Z$ for some positive integer $k$.} of
nilpotency class at most $p - 1$.

A typical application of the generalization we describe is the
situation where the group is a finite $p$-group of nilpotency class at
most $p$ and the Lie ring is a finite $p$-Lie ring of nilpotency class
at most $p$. Knowledge of either (the group or the Lie ring)
determines the other only up to isoclinism and not up to
isomorphism. Therefore, this correspondence is suited only for the
study of attributes (of groups or Lie rings) that are invariant under
isoclinism.

In cases where the original Lazard correspondence applies, it refines
the Lazard correspondence up to isoclinism: if a group and Lie ring
are in Lazard correspondence, then they are also in Lazard
correspondence up to isoclinism. The interesting case covered by our
correspondence is the case of finite $p$-groups of nilpotency class
exactly $p$ and finite $p$-Lie rings of nilpotency class exactly
$p$. The original Lazard correspondence no longer applies in this
situation,\footnote{There is a subtle distinction between the global
  and the $3$-local Lazard correspondence that we omit for the
  abstract, but describe in detail later} so our generalization
adds value.

\acknowledgments

I would like to thank George Glauberman, my research advisor, for help
over many years with the research underlying this thesis and for
detailed feedback on multiple drafts of the thesis. In addition I am
thankful to Peter May, my second adviser, for helping me out with a
number of the proofs and some of the underlying concepts. I would also
like to thank Professor Jonathan Alperin, who provided help in person,
via email, and by lending books and resources. In addition, I am
grateful to Shmuel Weinberger, the Department Chair, and Laurie Wail,
the Graduate Student Administrator, for being very accommodating in
helping me finish my thesis and arranging for financial support.

I am grateful to many mathematicians, both within and outside the
University of Chicago, for answering my many queries related to group
theory and mathematics. A partial list includes Max Horn of Justus
Liebig University Giessen, I. Martin Isaacs of the University of
Wisconsin-Madison, Ronald Solomon of Ohio State University, and Justin
Lynd, an assistant professor at Rutgers University. I am thankful to
Anton Geraschenko and StackExchange for collaborating to create the
mathematics question-and-answer website
MathOverflow,\footnote{\url{http://www.mathoverflow.net}} and to the
numerous mathematicians who volunteered their time to answering
questions of mine on MathOverflow, some of which have influenced
proofs that appear in this thesis. In addition, I am grateful to the
creators of GAP\footnote{Full list of authors at
  \url{http://www.gap-system.org/Contacts/People/authors.html}} for
providing a software that around with and better understand groups,
which are the central actors of my thesis.

I would also like to thank John Wiltshire-Gordon, who, while an
undergraduate at the University of Chicago, worked with me on related
problems. Some of the work we did together was the basis of Section
\ref{sec:bcuti}.

I also greatly enjoyed and benefited from reading the work of the many
mathematicians who are cited in this thesis, including Baer, Lazard,
Hall and Senior, Isaacs, Glauberman, Ellis, Khukhro, and many others
listed in the bibliography.

\mainmatter

\chapter{Introduction, outline, and preliminaries}

\input{thesis-intro.tex}

\chapter{Isoclinism and homoclinism: basic theory}

\input{thesis-isoclinism.tex}

\chapter{Extension theory}

\input{thesis-extension-theory.tex}

\chapter{Powering over sets of primes}

\input{thesis-powering.tex}

\chapter{Baer correspondence}

\input{thesis-baer.tex}

\chapter{The Malcev and Lazard correspondences}

\input{thesis-lazard-background.tex}

\chapter{Generalizing the Lazard correspondence to a correspondence up to isoclinism}

\input{thesis-main-proofs.tex}

\chapter{Applications and possible extensions}

\input{thesis-possible-extensions.tex}

\newpage

\begin{appendices}

\appendixpage
\addappheadtotoc

\chapter{Background material}

\input{thesis-appendix-part-1.tex}

\chapter{Supplementary proofs}

\input{thesis-appendix-part-2.tex}

\end{appendices}
\makebibliography

\begin{thebibliography}{99}

\singlespacing

\bibitem{AG98}{Jonathan L. Alperin and George Glauberman. {\em Limits
    of Abelian Subgroups of Finite $p$-Groups}. {\em Journal of
    Algebra}, Vol. 203 (1998), Page 533--566.}

\bibitem{Asch2004}{Michael Aschbacher. {\em The Status of the
    Classification of the Finite Simple Groups}. {\em Notices of the
    American Mathematical Society}, Vol. 51, No. 7
  (2004), Page 736--740. Available online at
  \url{http://www.ams.org/notices/200407/fea-aschbacher.pdf}}

\bibitem{Baer38} {Reinhold Baer. {\em Groups with Abelian Central
    Quotient Group}. {\em Transactions of the American Mathematical
    Society}, Vol. 44, No. 3 (November
  1938), Page 357--386. Available online at
  \url{http://www.jstor.org/stable/1989886}}

\bibitem{homologyofmultiplicativeLierings}{A. Bak, G. Donadze and
  N. Inassaridze. {\em Homology of multiplicative Lie rings}. {\em
    Journal of Pure and Applied Algebra}, Vol. 208
  (2007), Page 761--777. %Available online at
  \urlfake{http://www.math.uni-bielefeld.de/~bak/multliering.pdf}}

\bibitem{Baumslagunique}{Gilbert Baumslag. {\em Some aspects of groups
    with unique roots}. {\em Acta mathematica}, Vol. 104 (1960), Page
  217--303. %Available online at
%  \urlfake{http://eretrandre.org/rb/files/Baumslag1960_236.pdf}
}

\bibitem{BeylIsoclinisms}{F. R. Beyl. {\em Isoclinisms of group
    extensions and the Schur multiplicator}. {\em Groups St. Andrews
  1981}. {\em London Mathematical Society Lecture Note Series}, Vol. 71, Page
  169--185.}

\bibitem{BoyarchenkoSabitova}{Mitya Boyarchenko and Maria
  Sabitova. {\em The orbit method for profinite groups and a p-adic
    analogue of Brown's theorem}.  {\em Israel Journal of Math},
  Vol. 165 (2008), Page 67--91. Available on the ArXiV at
  \url{http://arxiv.org/abs/math/0608126}}

\bibitem{BrownLoday}{Ronald Brown and Jean-Louis Loday. {\em Van
    Kampen theorems for diagrams of spaces}. {\em Topology}, Vol. 26,
  No. 3 (1987), Page 311--335.} %Decomment after fixing hyperref:
                                    %Available online at
                                    %\urlfake{http://www-irma.u-strasbg.fr/~loday/PAPERS/87BrownLoday%28vanKampen%29.pdf}

\bibitem{Lazardeffective}{Serena Cicalo, Willem A. de Graaf and
  M. R. Vaughan-Lee. {\em An effective version of the Lazard
    correspondence}. {\em Journal of Algebra},  Vol. 352 (2012), Page
  430--450. Available online at
  \url{http://dx.doi.org/10.1016/j.jalgebra.2011.11.031}}

\bibitem{DummitFoote}{David S. Dummit and Richard M. Foote. {\em Abstract Algebra}. {\em John Wiley \& Sons} (2003). ISBN 0471433349}

\bibitem{Easterfield}{Timothy E. Easterfield. {\em The orders of
    products and commutators in prime-power groups}. {\em Cambridge
    Philosophical Society}, Vol. 36 (1940), Page 14--26. %Available
  %online at 
\urlfake{http://dx.doi.org/10.1017/S0305004100016960}}

\bibitem{EckmannHiltonStammbach}{Beno Eckmann, Peter J. Hilton, and
  Urs Stammbach. {\em On the homology theory of central group
    extensions: I - The commutator map and stem extensions}. {\em
    Commentarii Mathematici Helvetici}, Vol. 47, No. 1 (1972), Page 102--122. } %Decomment after fixing hyperref: Available online at
%  \urlfake{http://dx.doi.org/10.1007%2FBF02566792}

\bibitem{SchurmultiplierandLazard}{Bettina Eick, Max Horn, and Seiran
  Zandi. {\em Schur multipliers and the Lazard correspondence}. {\em
    Archiv der Mathematik}, Vol. 99 (2012), Page 217--226. Available
  online at \url{http://dx.doi.org/10.1007/s00013-012-0426-7}}

\bibitem{enumeratingpgroups}{Bettina Eick and E. A. O'Brien. {\em
    Enumerating p-groups}. {\em Journal of the Australian Mathematical
    Society}, Vol. 67 (1999), Page 191--205. %Available online at
  \urlfake{http://maths.auckland.ac.nz/~obrien/research/count-pgroups.pdf}}

\bibitem{Ellis87}{Graham Ellis. {\em The non-abelian tensor product of
    finite groups is finite}. {\em Journal of Algebra}, Vol. 111
  (1987), Page 203--205. Available online at
  \url{http://dx.doi.org/10.1016/0021-8693(87)90249-3}}

\bibitem{EllisLie}{Graham Ellis. {\em A non-abelian tensor product of
    Lie algebras}. {\em Glasgow Journal of Math}, Vol. 33 (1991), Page
  101--120.}

\bibitem{Ellis93}{Graham Ellis. {\em On five well-known commutator
    identities}. {\em Journal of the Australian Mathematical Society},
  Vol. 54 (1993), Page 1--19.}

\bibitem{Foguelinv}{Tuval Foguel and Abraham A. Ungar. {\em Involutory
    Decomposition of Groups Into Twisted Subgroups and
    Subgroups}. Available online at
  \url{http://paws.wcu.edu/tsfoguel/inv.pdf}}

\bibitem{Partialextensions}{George Glauberman. {\em A partial
    extension of Lazard's correspondence for finite $p$-groups}.{\em
    Groups, Geometry, and Dynamics}, Vol. 1, Issue 4 (2007). Available
  online at \url{http://dx.doi.org/10.4171/GGD/21}}

\bibitem{GG08}{George Glauberman, {\em Abelian subgroups of small
    index in finite $p$-groups}. {\em Journal of Group Theory}, Vol.
  8 (2005), Page 539--560. Available online at
  \url{http://dx.doi.org/10.1515/jgth.2005.8.5.539} }

\bibitem{GonzalezSanchez}{Jon Gonzalez-Sanchez. {\em Kirillov's orbit
    method for p-groups and pro-p-groups}. {\em Commutative Algebra}, Vol. 37, No. 12 (2009), Page 4476--4488.%Available online at \urlfake{http://www.uam.es/personal_pdi/ciencias/jgosanch/Publications/Kirillov_saturable_final.pdf}
}

\bibitem {HallSenior}{Marshall Hall and James Kuhn Senior. {\em The
    groups of order $2^n$ ($n \le 6$)}. %Reviewed on MathSciNet at
  \urlfake{http://www.ams.org/mathscinet-getitem?mr=168631}}

\bibitem {Hall37} {Philip Hall. {\em The classification of prime-power
    groups}. {\em Journal fur die reine und angewandte Mathematik},
  Vol. 69 (1937). Available online at
  \url{http://dx.doi.org/10.1515/crll.1940.182.130}}
  
\bibitem {Hekster}{N. S. Hekster. {\em On the structure of
    n-isoclinism classes of groups}, Vol. 40
  (1986), Page 63--85. Available online at
  \url{http://dx.doi.org/10.1016/0022-4049(86)90030-7}}

\bibitem{HiltonMislinRoitberg}{Peter Hilton, Guido Mislin, and Joe
  Roitberg. {\em Localization of Nilpotent Groups and Spaces}.
  {\em North-Holland Mathematics Studies} (55). {\em American Elsevier}
  (1975). ISBN 0444107762}

\bibitem{IsaacsCT}{I. Martin Isaacs. {\em Character Theory of Finite
    Groups}. {\em Dover} (2012). ISBN 0486680142}

\bibitem {Order128} {Rodney James, M. F. Newman, and
  E. A. O'Brien. {\em The groups of order 128}. {\em Journal of
    Algebra}, Vol. 129, No. 1
  (February 1990), Page 136--158. Available online at 
  \url{http://dx.doi.org/10.1016/0021-8693(90)90244-I}.}

\bibitem{Karpilovsky}{Gregory Karpilovsky. {\em The Schur
    Multiplier}. {\em Oxford University Press} (1987). ISBN 0198535546}

\bibitem{Khukhro}{Evgenii I. Khukhro. {\em $p$-Automorphisms of
    Finite $p$-groups}. {\em Cambridge University Press} (1998). ISBN
  052159717X}

\bibitem{Lazardsoriginal}{Michel Lazard. {\em Sur les groupes
    nilpotents et les anneaux de Lie}. {\em Annales scientifiques de
    l'E.N.S.} Third Series, Vol. 71, No. 2 (1954), Page 101--190.}

\bibitem{Baerinvariantsandisolosigms}{C. R. Leedham-Green and Susan
  McKay. {\em Baer-invariants, isologism, varietal laws and homology}.
  {\em Acta Mathematica}, Vol. 137, Issue 1 (1976), Page
  99--150. Available online at
  \url{http://dx.doi.org/10.1007/BF02392415}}

\bibitem{Mann99}{Avinoam Mann. {\em Some questions about
    $p$-groups}. {\em Journal of the Australian Mathematical Society},
  Vol. 67 (1999), Page 356--379. %Available online at \urlfake{http://journals.cambridge.org/download.php?file=%2FJAZ%2FJAZ1_67_03%2FS1446788700002068a.pdf}
}

\bibitem{enumeratingcharacters}{E. A. O'Brien and C. Voll. {\em
    Enumerating classes and characters of $p$-groups}. Available on the
  ArXiV at \url{http://arxiv.org/abs/1203.3050}}

\bibitem{ConciseII}{J. Peter May and Kate Ponto. {\em More Concise
    Algebraic Topology: Localization, Completion and Model
    Categories}. {\em University of Chicago Press} (2012). ISBN
  0226511782}

\bibitem{McDermottThesis}{Aidan McDermott. {\em The non-abelian tensor
    product of groups: computations and structural
    results}. Ph. D. thesis at National University of Ireland, Galway,
  under the supervision of Dr. Graham J. Ellis (February
  1998). Available online at
  \url{http://web.math.unifi.it/users/fumagal/articles/McDermott.pdf}}

\bibitem{Miller52} {Clair Miller. {\em The second homology group of a
    group; relations among commutators}. {\em Proceedings of the
    American Mathematical Society}, Vol. 3 (1952), Page
  588--595. Available online at
  \url{http://www.ams.org/journals/proc/1952-003-04/S0002-9939-1952-0049191-5/S0002-9939-1952-0049191-5.pdf}}

\bibitem{MilnorKTheory}{John Willard Milnor. {\em Introduction to
    algebraic K-theory}. {\em Annals of Mathematics Studies}
  (72), {\em Princeton University Press} (1971). ISBN 0691081018}

\bibitem{Moneyhun} {Kay Moneyhun. {\em Isoclinisms in Lie
    Algebras}. {\em Algebras Groups Geom.}, Vol. 11 (1994), Page 9--22.}

\bibitem {Order256}{E. A. O'Brien. {\em The groups of order
    256}. {\em Journal of Algebra}, Vol. 143 (1991), Page
  219--235. Available online at 
  \url{http://dx.doi.org/10.1016/0021-8693(91)90261-6}}

\bibitem{Moghaddametal}{Foroud Parvaneh, Mohammad Reza R. Moghaddam,
  A. Khaksar. {\em Some Properties of $n$-Isoclinism in Lie
    Algebras}. {\em Italian Journal of Pure and Applied
    Mathematics}, Vol. 28 (2011), Page 165--176. %Available online at
%\urlfake{http://ijpam.uniud.it/online_issue/201128/18-ParvaneMoghadKhaksar%20n-isoclinism.pdf}
}

\bibitem{Pilgrim}{Donald H. Pilgrim. {\em Engel conditions on
    groups}. {\em Proceedings of the Iowa Academy of Sciences}, Vol.
  71 (1964), Page 377--383. %Reviewed on MathSciNet at
  %\urlfake{http://www.ams.org/mathscinet-getitem?mr=0232851}
}

\bibitem{Plotkin}{B. I. Plotkin. {\em On some criteria of locally
    nilpotent groups}. {\em Uspekhi Mat. Nauk}, 9:3 (61) (1954), Page 181--186. %Available online at
  \urlfake{http://www.mathnet.ru/php/archive.phtml?wshow=paper&jrnid=rm&paperid=8095&option_lang=eng}}

\bibitem{RotmanGT}{Joseph J. Rotman. {\em An Introduction to the
    Theory of Groups}. {\em Springer Graduate Texts in Mathematics}
  (1999). ISBN 0387942858}

\bibitem{Stallings}{John Stallings. {\em Homology and central series
    of groups}. {\em Journal of Algebra}, Vol. 2 (1965), Page 170--181}.

\bibitem {SuzukiII}{Michio Suzuki. {\em Group Theory II}. {\em
    Springer} (1986). ISBN 0387109161}

\bibitem{vsv}{Veeravalli S. Varadarajan. {\em The Lie group-Lie
    algebra correspondence}. Lecture notes on Lie theory, Chapter
  9. %Available online at
  \urlfake{http://www.math.ucla.edu/~vsv/liegroups2007/9.pdf}}

\bibitem{Weibel}{Charles Weibel. {\em An Introduction to Homological
    Algebra}. {\em Cambridge University Press} (1995). ISBN
  0521559871}

\bibitem{Weigel}{Thomas S. Weigel. {\em Exp and log functors for the
    categories of powerful $p$-central groups and Lie
    algebras}. Habilitationsschrift (Ph.D. thesis) at Freiburg im
  Breisgau (1994).}

\end{thebibliography}

\end{document}
